### 2007-07-10 bsis_paper/5-BSIS.tex
BSIS chapter from dissertation.
git-svn-id: file:///home/rblake/svn/pamg/trunk@8 c7f3fe13-8c30-0410-9127-9a9c42fe009c

\chapter{Bucket Sorted Independent Sets}
\label{chap:BSIS}

\section{Introduction}
Chapters~\ref{chap:CGS} and~\ref{chap:color-based} demonstrate the
behavior of independent set-based coarse-grid selection
algorithms. One common design property among the selection algorithms
is a routine to search for vertices to become
$C$-points. Additionally, an update in the weights of vertices as the
coarse-grid selection often follows. In this chapter, new theory and
algorithms to decrease the computational cost associated with the
search and weight update procedures used in coarse-grid selection are
presented.

\section{CLJP-c}
Recall that CLJP-c colors the graph of $S$ before selecting
$C$-points, and the colors are used as one component of the vertex
weights. As a result, the structure of the coarse grids selected is
improved. More formally, the use of graph coloring provides the
following important result.
\begin{theorem}
\label{5:theorem:CLJPc-neighbor-independence}
For all pairs of vertices $i$ and $j \in \mathcal{N}_i$ CLJP-c
guarantees $w_i \ne w_j$.
\end{theorem}
\begin{proof}
Assume two adjacent vertices $i$ and $j$ have the same weight. That is,
$|S_i^T| = |S_j^T|$ and the weight augmentation provided through
coloring is the same for $i$ and $j$. The graph of $S$, however, is
colored such that $i$ and $j$ are different colors for all $j \in
\mathcal{N}_i$, so a contradiction is reached.
\end{proof}
Theorem~\ref{5:theorem:CLJPc-neighbor-independence} establishes that
all adjacent vertices have different weights in CLJP-c, which is not
guaranteed in CLJP, although is unlikely to occur. The following
corollaries are a result of
Theorem~\ref{5:theorem:CLJPc-neighbor-independence}.
\begin{corollary}
Any set of vertices in the graph of $S$ that share the same weight is
an independent set.
\end{corollary}
\begin{corollary}
\label{5:cor:max-weight-indep-set}
The set of vertices with the largest weight in the graph of $S$ form
an independent set satisfying~\ref{5:eqn:D-conditions}. That is, each
vertex in that independent set has a uniquely maximal weight in its
neighborhood.
\end{corollary}
The first corollary states that independent sets can be selected in
the graph simply by selecting sets of vertices with the same
weight. Corollary~\ref{5:cor:max-weight-indep-set} refines this
observation to a subset of vertices guaranteed to satisfy the
selection criterion. In particular, it shows it is possible to build
the coarse grid by selecting vertices with the maximum weight,
updating weights, selecting the next set of vertices with maximum
weight, and so on. This approach is taken by the algorithm developed
in Section~\ref{5:sec:contribution}. It is proven in
Section~\ref{5:sec:cg-invariance} that despite the difference in
approach, the resulting coarse grid is the same as that selected by
CLJP-c.

\section{Coarse-Grid Selection Search and Weight Update}
\label{5:sec:s-and-w-update}
CLJP and its descendents select an independent set $D$ in each
iteration. Stated originally in Section~\ref{3:sec:CLJP}, the
condition for selecting $D$ is that all vertices $i \in D$ must
satisfy
\begin{equation}
\label{5:eqn:D-conditions}
w_i > w_j\;\textrm{for all } j \in \mathcal{N}_i.
\end{equation}
CLJP and CLJP-c rely on a search routine to locate locally maximal
weights and on a weight update routine to modify weights of vertices
connected to new $C$-points.

The algorithms for searching and updating vertex weights in CLJP in
detail are examined in this section. In particular, the impact of
using a sparse matrix format on the coarsening procedure. The
pseudo-code below assumes the software uses a compressed sparse row
(CSR)~\cite{saadBook} matrix format or other similar format, which are
common matrix formats in numerical software. CSR provides low memory
costs for storing sparse matrices and provides efficient access to the
nonzeros in a row. Accessing the nonzeros in a column is an expensive
operation in this format and strongly influences the weight update
routine in CLJP-style algorithms because $S$ is, in general, not
symmetric.

The search step in CLJP is implemented as shown in
Algorithm~\ref{5:alg:search}.
\begin{algorithm}[h]
\caption{Coarse-Grid Selection Graph Search}
\label{5:alg:search}
\textsc{Search-Graph}$(S,\, C,\, F)$ \{
\begin{algorithmic}[1]
\STATE $D \leftarrow \emptyset$
\FORALL{$i \notin (C \cup F)$}
\IF{$w_i > w_j,\,\forall j \in \mathcal{N}_i$}
\STATE $D \leftarrow D \cup \{i\}$
\ENDIF
\ENDFOR
\RETURN $D$
\end{algorithmic}
\}
\end{algorithm}
In the first iteration, Line~3 is run $2|E|$ times. The total cost of
search in constructing the coarse grid depends on the number of
iterations needed. Even in the best case of $\Omega(E)$ time, the cost
is significant when the graph contains large numbers of edges, as
usually happens on the lower levels in the grid hierarchy
(see~\cite{alber-PCGS} for examples). In the next section, a new
technique is introduced for conducting the search in coarse-grid
selection algorithms independent of the number of edges in the graph.

Pseudo-code for the weight update in CLJP is shown in Algorithm~\ref{5:alg:setup}.
\begin{algorithm}[h]
\caption{CLJP Weight Update for CSR Matrix}
\label{5:alg:setup}
\textsc{Update-Weights}$(S,\, D,\, C,\, F,\, w)$ \{
\begin{algorithmic}[1]
\FORALL{$d \in D$}
\FORALL{$i \in S_d$}
\STATE $w_i \leftarrow w_i - 1$ \COMMENT{see
  Figure~\ref{3:fig:CLJP-update-heuristics} left}
\STATE $S_d = S_d \setminus \{i\}$ \COMMENT{removing edge from graph}
\ENDFOR
\ENDFOR
\FORALL{$i \notin (C \cup F)$}
\FORALL{$k$ originally in $S_i$}
\IF{$k \in D$}
\STATE mark $k$ NEW-C-POINT
\ENDIF
\ENDFOR
\FORALL{$j \in S_i$}
\IF{$j \notin D$}
\FORALL{$k \in S_j$}
\IF[$i$ and $j$ mutually influenced by $k$]{$k$ is marked NEW-C-POINT}
\STATE $w_j \leftarrow w_j - 1$ \COMMENT{see
  Figure~\ref{3:fig:CLJP-update-heuristics} right}
\STATE $S_i = S_i \setminus \{j\}$ \COMMENT{remove edge from $j$ to $i$}
\ENDIF
\ENDFOR
\ENDIF
\ENDFOR
\FORALL{$k$ originally in $S_i$}
\IF{$k \in D$}
\STATE unmark $k$
\STATE $S_i = S_i \setminus \{k\}$ \COMMENT{remove edge from $k$ to $i$,
  if present}
\ENDIF
\ENDFOR
\ENDFOR
\end{algorithmic}
\}
\end{algorithm}
The level of complication in this update routine is due to the CSR
format and the need to find vertices strongly influenced by new
$C$-points. When a new $C$-point $k$ is selected, the first type of
weight update is trivial since determining the vertices in $S_k$ is
inexpensive. The second type of update is more expensive since the
vertices influenced by $k$ are difficult to determine in a CSR
format. The update requires a search of many vertices $i$ and all of
their strong influencing neighbors $j$. The routine then searches
strongly influencing $j$ to determine if any $k \in D$ strongly
influences both $i$ and $j$. The cost increases dramatically as the
density of $S$ increases. Large operator complexities have a
disproportionately large impact on coarse-grid selection run time. In
the next section, a modified update routine to compliment the new
search technique is introduced.

\section{Bucket Sorted Independent Set Selection}
\label{5:sec:contribution}
In this section, new techniques for searching the graph of $S$ for new
$C$-points and subsequently updating the weights of remaining vertices
are developed. The new algorithm is labeled Bucket Sorted Independent
Sets (BSIS) to reflect the data structure used.

Like CLJP-c, BSIS depends on graph coloring, but utilizes modified
routines for search and weight update. Furthermore, rather than
applying the color information to augment vertex weights, BSIS uses
the colors in a bucket data structure. Once initialized, this data
structure selects independent sets, which satisfy the conditions in
(\ref{5:eqn:D-conditions}), in constant time.

\subsection{Coarse Grid Invariance}
\label{5:sec:cg-invariance}
Theory is developed in this section to prove coarse-grid selection
invariance in general independent set-based algorithms. The algorithms
considered thus far select independent sets using
(\ref{5:eqn:D-conditions}), meaning $i$ is eligible to be in $D$ if
its weight is larger than the weights of vertices in its neighborhood
$\mathcal{N}_i$. Recall the neighborhood of $i$
(Definition~\ref{3:def:neighborhood}) is the set of vertices strongly
influenced by $i$ or strongly influencing $i$. Algorithms relying on
different and larger neighborhoods, such as \emph{distance-$d$
neighborhoods}, are conceivable.
\begin{definition}
The \emph{distance-$d$ neighborhood} of $i$, denoted
$\mathcal{N}_i^d$, is the set of vertices within $d$ hops of $i$ in
the symmetrized strength matrix, excluding $i$. That is,
$\mathcal{N}_i^d = \left[\bigcup_{j \in
\mathcal{N}_i^{d-1}}\left(\mathcal{N}_j \cup \{j\}\right)\cup
\mathcal{N}_i\right] \setminus \{i\}$, where $d > 0$ and
$\mathcal{N}_i^0 = \emptyset$.
\end{definition}

For a general independent set-based coarse-grid selection algorithm, a
vertex $i$ is eligible to be added to $D$ if
\begin{equation}
\label{5:eqn:general-selection}
w_i > w_j\;\textrm{for all } j \in \mathcal{N}_i^s,
\end{equation}
where $\mathcal{N}_i^s$ is the \emph{selection neighborhood} of
$i$. Although the distance-$d$ neighborhood is a sensible choice for
the selection neighborhood, this discussion is not limited to such
cases. It is assumed, however, the matrix formed by $\mathcal{N}_*^s$
(i.e., the selection sets for all vertices) is symmetric. This is
equivalent to stating if $j \in \mathcal{N}_i^s$, then $i \in
\mathcal{N}_j^s$ for all $i$ and $j$ in the vertex set. Furthermore,
this discussion assumes the weight of vertex $i$ is only decremented
following weight updates and is never modified due to the assignment
of a vertex $j \notin \mathcal{N}_i^s$ to the $C$-point set.

Given a symmetric, but otherwise arbitrary, set of selection
neighborhoods, the set of vertices potentially able to affect the
vertex weight of $i$ or $j \in \mathcal{N}_i^s$ is the \emph{extended
selection neighborhood}.
\begin{definition}
The \emph{extended selection neighborhood} of $i$, denoted
$\mathcal{N}_i^{2s}$, is the union of the selection neighborhood of
$i$ with the selection neighborhoods of the vertices in
$\mathcal{N}_i^s$, excluding $i$. That is, $\mathcal{N}_i^{2s} =
\left[\left(\bigcup_{j \in \mathcal{N}_i^s}\mathcal{N}_j^s\right)\cup
\mathcal{N}_i^s\right] \setminus \{i\}$.
\end{definition}

When a vertex $i$ satisfies the generalized selection
condition~(\ref{5:eqn:general-selection}), no other $C$-point
assignments affect $w_i$. This is formalized below.
\begin{lemma}
\label{5:lemma:C-lock}
If (\ref{5:eqn:general-selection}) is satisfied for vertex $i$, then
$i$ must be the next vertex in $\{i\} \cup \mathcal{N}_i^s$ to become
a $C$-point, regardless of any other selections and updates made in
the graph.
\end{lemma}
\begin{proof}
The satisfaction of (\ref{5:eqn:general-selection}) means $w_i > w_j$
for all $j \in \mathcal{N}_i^s$. For $i$ to not become a $C$-point,
its weight must become smaller than some $j \in \mathcal{N}_i^s$. The
weight of $i$, however, is not decremented unless some $j \in
\mathcal{N}_i^s$ satisfies~(\ref{5:eqn:general-selection}) and becomes
a $C$-point, which is impossible until after $i$ is assigned to the
$C$-point set.
\end{proof}

To demonstrate all algorithms using the same selection neighborhood
and update rules select identical coarse grids, a proof with an
inductive argument is presented below. The base case is provided by
the first set of vertices satisfying the general selection conditions.
\begin{definition}
Let $D_0$ be the set of all vertices satisfying
(\ref{5:eqn:general-selection}) in the first iteration of coarse-grid
selection.
\end{definition}

Vertices in $D_0$ are destined to become $C$-points regardless of the
algorithm used to build coarse grids. In the independent set-based
algorithms discussed in Chapters~\ref{chap:CGS}
and~\ref{chap:color-based}, all $D_0$ vertices become $C$-points in
the first iteration. Any algorithm constructed, however, eventually
selects all $D_0$ vertices as $C$-points, as proven in the following
lemma.
\begin{lemma}
\label{5:lemma:D0_invariant}
Given the same selection neighborhoods and update heuristics, all
algorithms select vertices in $D_0$ as $C$-points.
\end{lemma}
\begin{proof}
The proof follows from Lemma~\ref{5:lemma:C-lock}. All vertices in
$D_0$ satisfy the selection condition~(\ref{5:eqn:general-selection}),
so the assignment of any other $C$-point has no effect on the weight
of a $D_0$ vertex.
\end{proof}

Lemma~\ref{5:lemma:D0_invariant} states that any coarse-grid selection
method using the general selection condition invariably selects $D_0$
vertices as $C$-points. This result is used in the proof of the next
theorem.

\begin{theorem}
\label{5:theorem:invariance}
All independent set-based coarse-grid selection algorithms given the
same initial weights and using the same selection neighborhood,
selection criterion based on~(\ref{5:eqn:general-selection}), and
weight update heuristics as described above select identical coarse
grids.
\end{theorem}
\begin{proof}
Let $c$ be the vertices in $\mathcal{N}_i^{2s}$ satisfying the
conditions for $D$ in some arbitrary iteration. Suppose assigning
vertices $c_1 \subset c$ to the $C$-point set leads to $w_i > w_j$ for
all $j \in \mathcal{N}_i^s$. Also suppose assigning $c_2 \subset c$,
$c_1 \ne c_2$, to the $C$-point set leads to the existence of some $j
\in \mathcal{N}_i^s$ such that $w_j > w_k$ for all $k \in
\mathcal{N}_j^s$.

For both conditions to be true, one or both of the following cases
must be satisfied.
\begin{enumerate}
\item The value of $w_i$ is smaller when $c_2$ is added to the
  $C$-point set than when $c_1$ is added. For this case, it must be
  true that $|c_2 \cap \mathcal{N}_i^s| > |c_1 \cap \mathcal{N}_i^s|$.
\item The value of $w_j$ is larger when $c_2$ is added to the
  $C$-point set than when $c_1$ is added. For this case, it must be
  true that $|c_2 \cap \mathcal{N}_j^s| < |c_1 \cap
  \mathcal{N}_j^s|$.
\end{enumerate}

Case~1 creates a contradiction. If $|c_2 \cap \mathcal{N}_i^s| > |c_1
\cap \mathcal{N}_i^s|$, then $(c \setminus c_1) \cap \mathcal{N}_i^s
\ne \emptyset$. Following the assignment of the vertices in $c_1$ to
the $C$-point set, there remains some $k \in \mathcal{N}_i^s$ that is
also in $c$. Therefore, $w_k > w_i$, contradicting the first assumed
condition.

Case~2 is similarly impossible. If $|c_2 \cap \mathcal{N}_j^s| < |c_1
\cap \mathcal{N}_j^s|$, then $(c \setminus c_2) \cap \mathcal{N}_j^s
\ne \emptyset$. Following the assignment of the vertices in $c_2$ to
the $C$-point set, there remains some $k \in \mathcal{N}_j^s$ that is
also in $c$. Therefore, $w_k > w_j$, contradicting the second assumed
condition.

Both cases are impossible, so the order of $C$-point selection within
the selection neighborhood of each vertex is invariant. Combined with
Lemma~\ref{5:lemma:D0_invariant} as the base case, this proves by
induction the invariance of coarse-grid selection for all algorithms
using identical selection conditions.
\end{proof}

\begin{remark}
CLJP and CLJP-c use the distance-one neighborhood as the selection
neighborhood, (\ref{5:eqn:D-conditions}) as the selection criterion,
and the weight update heuristics in
Algorithm~\ref{3:alg:CLJP-update-heuristics}. Given the same initial
weights, all algorithms based on the parameters utilized by CLJP and
CLJP-c select identical coarse grids.
\end{remark}

Theorem~\ref{5:theorem:invariance} is an important result about the
nature of coarse grids selected by general independent set-based
algorithms. This information enables the design and implementation of
new algorithms that yield identical coarse grids using different and
possibly more efficient techniques.

\subsection{Bucket Sorted Independent Sets Algorithm}
The BSIS algorithm creates a bucket data structure enabling a search
for new $C$-points without individually scanning each vertex and
associated edges. Figure~\ref{5:fig:bucket-algorithm} illustrates the
bucket data structure. The number of buckets in the data structure is
$\max_{i \in S}|S_i^T|$ times the number of colors in the graph. That
is, each possible weight in $S$ has its own bucket. The vertices are
distributed to the appropriate buckets during the setup of the
coarse-grid selection algorithm, where the bucket of a newly placed
vertex depends on the number of vertices it strongly influences and
its color. For example, Vertex~14 strongly influences six vertices and
is black. Therefore, it is placed into the black bucket in the sixth
group of buckets. More notably, the vertices in a bucket form an
independent set (e.g., vertices 4, 16, and 21).

In each iteration, the non-empty bucket with largest weight forms $D$
(see Corollary~\ref{5:cor:max-weight-indep-set}). These vertices are
assigned to the $C$-point set and removed from the data
structure. Vertex weight updates lead to corresponding updates to the
data structure, and new $F$-points are removed from the data
structure. These operations continue until all buckets are empty, at
which point the coarse-grid selection is
complete. Algorithms~\ref{5:alg:BSIS-setup},
\ref{5:alg:BSIS-indep-set-select}, and~\ref{5:alg:BSIS-weight-update}
outline the operations discussed above.
\begin{algorithm}[!h]
\caption{BSIS Data Structure Setup}
\label{5:alg:BSIS-setup}
\textsc{BSIS-Setup}$(S)$ \{
\begin{algorithmic}[1]
\FORALL{$i \in V$}
\STATE $bucketID \leftarrow (w_i-1)\cdot numColors + color_i$
\STATE $bucket[bucketID].\textsc{Insert}(i)$
\ENDFOR
\end{algorithmic}
\}
\end{algorithm}
\begin{algorithm}[!h]
\caption{Independent Set Selection}
\label{5:alg:BSIS-indep-set-select}
\textsc{BSIS-Independent-Set-Selection}$(S)$ \{
\begin{algorithmic}[1]
\RETURN non-empty $bucket$ with largest $bucketID$
\end{algorithmic}
\}
\end{algorithm}
\begin{algorithm}[!h]
\caption{BSIS Weight Update}
\label{5:alg:BSIS-weight-update}
\textsc{BSIS-Weight-Update}$(S)$ \{
\begin{algorithmic}[1]
\STATE $bucketID \leftarrow (w_i-1)\cdot numColors + color_i$
\STATE $bucket[bucketID].\textsc{Remove}(i)$
\STATE $bucket[bucketID - numColors].\textsc{Insert}(i)$
\end{algorithmic}
\}
\end{algorithm}
\begin{figure}
  \begin{center}
    %\includegraphics[width=0.8\textwidth]{images/5/bucket-algorithm.eps}
    \includegraphics[width=0.98\textwidth]{images/5/BSIS/initial.eps}
    \caption{The BSIS data structure.}
    \label{5:fig:bucket-algorithm}
  \end{center}
\end{figure}

Figure~\ref{5:fig:bucket-algorithm-cp1} illustrates the graph and data
structure following the first iteration of the algorithm. Vertex~10
has become a $C$-point and its neighbors weights have been
updated. Vertices assigned to $F$ or $C$ are removed from the data
structure and other affected vertices are moved to new locations in
the data structure. Such vertices are highlighted in red.
\begin{figure}
  \begin{center}
    \includegraphics[width=0.98\textwidth]{images/5/BSIS/cp1.eps}
    \caption{The BSIS data structure after selecting the first
    $C$-point (Vertex 10). The weights of neighbors of the new
    $C$-point are updated. Some neighbors become $F$-points and are
    removed from the data structure. Vertices removed from the data
    structure are highlighted with a red ring in the graph, while
    other neighbors are moved to new locations in the data structure
    and are highlighted (to help in reading the figure) in the data
    structure with a red box.}
    \label{5:fig:bucket-algorithm-cp1}
  \end{center}
\end{figure}

The weight update routine described in
Section~\ref{5:sec:s-and-w-update} is very expensive in this context
because some iterations of BSIS select few $C$-points. For a graph
with a large number of colors, BSIS may execute dozens or hundreds of
low-cost iterations to select a coarse grid. Recall the weight update
routine described loops through all unassigned vertices each time it
is called, so when it is run by BSIS, work is done on many unaffected
vertices, which is computationally inefficient.

The largest factor in the cost of the weight update results from
searching for the second type of weight update in
Algorithm~\ref{5:alg:setup}, which is done by looping through all
unassigned vertices since a new $C$-point cannot easily determine
which vertices it strongly influences in a CSR matrix. It is less
expensive in this situation to construct the transpose of $S$ than to
search the entire graph in each iteration. In $S^T$, a $C$-point
quickly determines which vertices it influences and ``paints''
them. The algorithm then loops through all painted vertices and
determines if any are neighbors. This is a simple solution that has a
dramatic effect on the performance of BSIS, although the update cost
remains approximately equivalent to the cost in
CLJP-c. Chapter~\ref{chap:conclusions} describes plans for continued
investigation into decreasing the cost of weight update in coarsening
algorithms.

\section{Weight Update Aggregation}
\label{5:sec:agg}
Whenever a vertex weight is updated, BSIS moves the corresponding
vertex to a new location in the data structure. During the selection
of the coarse grid, the cost of the updates to the data structure is
non-trivial and, as shown in this section, unnecessary.

Only one bucket in the data is touched during the $C$-point selection
step: the largest weight non-empty bucket in the data structure. Other
buckets are subsequently affected by the weight updates resulting from
new $C$-point assignments. A different approach is possible, however,
since the only bucket that must contain the correct vertices is the
one from which $C$-points are selected.

To save cost, we suggest a lazy approach based on aggregating the cost
of the updating vertex weights. Rather than investing computation into
maintaining accuracy in the data structure, a less expensive mechanism
to test if a vertex is in the correct location is provided. When a
weight is updated, the vertex is not moved until it is found in the
bucket being used as the new independent set $D$.

Figure~\ref{5:fig:cp1-agg} depicts the data structure after the first
set of $C$-points is selected.
\begin{figure}
  \begin{center}
    \includegraphics[width=0.98\textwidth]{images/5/BSIS/cp1-agg.eps}
    \caption{The BSIS data structure after selecting the first
    $C$-point (Vertex 10) with aggregate weight updates.}
    \label{5:fig:cp1-agg}
  \end{center}
\end{figure}
Rather than moving vertices to new buckets, the method now keeps them
in the same location and only moves them when necessary. As shown in
Section~\ref{5:sec:experiments}, aggregation of the weight updates
leads to significant savings in computational cost.

\section{Experimental Results}
\label{5:sec:experiments}
To demonstrate BSIS, the algorithm is compared with CLJP-c. The test
problem is the 3D 7-point Laplacian on a structured grid:
\begin{eqnarray}
\label{5:eqn:Lap} -\Delta u & = & 0 \quad \textrm{on } \Omega \qquad
(\Omega = (0, 1)^3),\\
\nonumber u & = & 0 \quad \textrm{on } \partial \Omega.
\end{eqnarray}
A 7-point Laplacian is selected since it is a common initial test
problem and structured problems often lead to the largest operator
complexities for coarsening algorithms satisfying heuristic H1. By
creating larger operator complexities, the algorithms are forced to
traverse more edges in the graph, leading to more work.

Timing data for the selection of all coarse grids in the hierarchy is
reported. This time includes the cost for all parts of the algorithms,
including the graph coloring phase. AMG solve phase information is not
reported since the algorithms produce identical coarse grids and since
information on solve phase performance for AMG with CLJP-c is
documented in~\cite{alber-cljpc, alber-PCGS}.

The smallest problem is a $30 \times 30 \times 30$ grid. Subsequent
problems are grids of size $60^3$, $90^3$, up to $210^3$. The largest
problem is 343 times larger than the smallest problem and contains
more than nine million degrees of freedom.

Results for the experiment are presented in
Figure~\ref{5:fig:results-full}.
\begin{figure}
  \begin{center}
    \includegraphics[width=0.75\textwidth]{images/5/agg-no-scaling.eps}
    \caption{Coarse-grid selection times using BSIS with aggregate
    weight update, standard BSIS, and CLJP-c on the 7-point
    Laplacian.}
    \label{5:fig:results-full}
  \end{center}
\end{figure}
BSIS completes coarse-grid construction in less time than CLJP-c in
every case, and BSIS with aggregate weight update performs
significantly better than standard BSIS. For the largest problems BSIS
is approximately 17\% cheaper than CLJP-c. BSIS with aggregate weight
updates is 23\% cheaper on the largest problems. The benefit is
magnified, relative to CLJP-c, for the smaller problems.

Experiment (\ref{5:eqn:Lap}) demonstrates the effectiveness and
competitiveness of the bucket technique. Increased efficiency for the
BSIS algorithm is anticipated through further research. Furthermore,
the methods and concepts in this research are also applicable to other
coarsening algorithms and possibly to other elements in the AMG setup
phase.

\section{BSIS Variants}
The ideas developed in this chapter are applicable to other coarsening
algorithms that utilize a graph coloring step, such as color-based
methods designed to satisfy heuristic H1$^\prime$. The PMIS-c1
algorithm is expected to benefit from BSIS and is expected to perform
naturally in parallel since vertex weights are not updated in PMIS.

\section{Parallelizing BSIS}
Using BSIS in a parallel algorithm presents challenges because the
parallelism in BSIS is very fine-grained. Its elegance and potential
to greatly improve the efficiency of coarse-grid selection motivates
the development of parallel algorithms incorporating BSIS. Several
alternatives are explored in this section.

The first idea is called the boundary painting method and takes
advantage of the invariance between coarse grids selected by BSIS and
CLJP-c. The idea is to use BSIS to select as many of the interior
$C$-points as possible before doing any communication with neighboring
processors. All vertices belonging to a processor are colored and
inserted into the BSIS data structure. The processor boundary
vertices, however, are ``painted''. If a painted vertex is in a set
$D$ in some iteration, then the vertex is not added to $C$. It is
instead removed from the data structure and its on-processor neighbors
are also painted. Figure~\ref{5:fig:cp1-parallel} illustrates the
first iteration of the painted boundary method with weight update
aggregation. The data structure shown is for the left domain.
\begin{figure}
  \begin{center}
    \includegraphics[width=0.98\textwidth]{images/5/BSIS/cp1p.eps}
    \caption{The painted boundary method with aggregate weight
    updates following one iteration. The data structure is for the
    left domain. Painted vertices are marked with a green ring in the
    graph and a green box in the data structure. Vertices in the data
    structure that are half green and half red are painted and also
    have had their weights updated.}
    \label{5:fig:cp1-parallel}
  \end{center}
\end{figure}
The first iteration selects a new $C$-point, but does not select any
painted vertices. In the second iteration, Vertex~22 is selected, but
is already painted. Therefore, on-processor neighbors of Vertex~22 are
also painted (see Figure~\ref{5:fig:cp2-parallel}).
\begin{figure}
  \begin{center}
    \includegraphics[width=0.98\textwidth]{images/5/BSIS/cp2p.eps}
    \caption{The painted boundary method with aggregate weight
    updates following the second iteration. In this iteration a
    painted vertex was selected, leading to the painting of its
    on-processor neighbors.}
    \label{5:fig:cp2-parallel}
  \end{center}
\end{figure}
The method finishes when the data structure is emptied of all
vertices. The product is now three disjoint sets: $C$ and $F$, as
usual, but also a set of painted vertices. The painted vertices are
the vertices that cannot be assigned to the $F$ or $C$ set without
communication with another processor. The information is provided to
CLJP-c, which handles the parallel portion of the algorithm.

The painted boundary approach is ideal given large numbers of interior
vertices, which can be guaranteed on most problems for the fine grid.
A side-effect of H1-based coarsening algorithms, however, is the
creation of denser graphs on coarse levels. One solution is to use the
painted boundary method on fine levels and then switch to CLJP-c
further along in the coarsening process. The issue is a smaller
concern for H1$^\prime$-based coarsening using BSIS since these
methods produce small operator complexities that do not grow as the
size of the problem is increased. In all cases, however, the number of
processor boundary vertices relative to the number of interior
vertices increases as the number of unknowns per processor decreases
(e.g., on coarse levels). A few techniques may be applicable in this
situation. The easiest solution is to simply use CLJP-c, or some other
coarsening algorithm, on the coarser levels where few vertices are
found. Although BSIS does not decrease in cost at this point, the
total cost on the levels when BSIS is not used is low due to low
complexity of coarse grids. A second approach is the application of a
dynamic load balancing algorithm to increase the number of vertices on
a processor (and, thus, decrease communication costs). If the number
of vertices per processor per level is maintained at a high enough
level, BSIS is still valuable on coarse grids. A third option is to
replicate the operator matrix on processors, which leads to processors
doing more of the same work, but by avoiding communication. The second
and third ideas are similar in nature and both involve using dynamic
load balancing techniques~\cite{DevineBomanKarypisPP04, CybDLB,
deCoughnyLB, SchloegelLB, ZoltanParHyp06ipdps, ZoltanParHypRepart07}.

### 2007-07-31 bsis_paper/paper.tex
started paper.tex with place holders for the images in the images directory
git-svn-id: file:///home/rblake/svn/pamg/trunk@9 c7f3fe13-8c30-0410-9127-9a9c42fe009c

--- 2007-07-10 bsis_paper/5-BSIS.tex
@@
-\chapter{Bucket Sorted Independent Sets}
-\label{chap:BSIS}
+\documentclass{elsart}
+
+% Use the option doublespacing or reviewcopy to obtain double line spacing
+% \documentclass[doublespacing]{elsart}
+
+% if you use PostScript figures in your article
+% use the graphics package for simple commands
+% \usepackage{graphics}
+% or use the graphicx package for more complicated commands
+\usepackage[draft]{graphicx}
+% or use the epsfig package if you prefer to use the old commands
+% \usepackage{epsfig}
+
+\usepackage{amssymb,listings}
+
+%% THE ENVIRONMENT \begin{proof} ... \end{proof} PRODUCES AN
+%% END-OF-PROOF SIGN.
+\newenvironment{proof}{\begin{pf}}{\qed\end{pf}}
+%% THE ENVIRONMENT \begin{proof*} ... \end{proof*} SUPPORTS ADDING
+%% INFORMATION SUCH AS  'Proof (Sketch)' AS THE LABEL FOR THE PROOF.
+\newenvironment{proof*}{\begin{pf*}}{\qed\end{pf*}}
+
+% The lineno packages adds line numbers. Start line numbering with
+% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
+% for the whole article with \linenumbers.
+% \usepackage{lineno}
+% \linenumbers
+
+\usepackage{listings}
+\lstset{mathescape,
+        escapechar=\%,
+        numbers=left, 
+        numberstyle=\tiny, 
+        stepnumber=1, 
+        numbersep=5pt,
+        frame=tb,
+        xleftmargin=5mm,
+        framexleftmargin=5mm,
+        language=Matlab,
+        belowcaptionskip=5pt}
+\renewcommand{\lstlistingname}{Algorithm}
+\usepackage{caption}
+\captionsetup[lstlisting]{justification=raggedright}
+
+\begin{document}
+\begin{frontmatter}
+
+% Title, authors and addresses
+
+% use the thanksref command within \title, \author or \address for footnotes;
+% use the corauthref command within \author for corresponding author footnotes;
+% use the ead command for the email address,
+% and the form \ead[url] for the home page:
+\title{Bucket-Sorted Independent Sets for Algebraic Multigrid}
+\author{David Alber\corauthref{cor}},
+\corauth[cor]{Corresponding author. Address: 201 N. Goodwin Ave., Siebel
+Center for Computer Science, Urbana, IL 61801}
+\ead{dalber@uiuc.edu}
+\ead[url]{http://www.cs.uiuc.edu/homes/dalber}
+\author{Luke Olson}
+\address{Department of Computer Science\\ University of Illinois at Urbana-Champaign}
+\ead{dalber@uiuc.edu}
+\ead[url]{http://www.cs.uiuc.edu/homes/lukeo}
+% \thanks[label2]{}
+% \address{Address\thanksref{label3}}
+% \thanks[label3]{}
+% use optional labels to link authors explicitly to addresses:
+% \author[label1,label2]{}
+% \address[label2]{}
+
+\begin{abstract}
+% Text of abstract
+\end{abstract}
+
+\begin{keyword}
+multigrid \sep parallel algorithms
+% keywords here, in the form: keyword \sep keyword
+
+% PACS codes here, in the form: \PACS code \sep code
+% \PACS 
+\MSC 65F10 \sep 65Y05 \sep 05C69
+\end{keyword}
+\end{frontmatter}
+
+% main text
 
@@
 following important result.
-\begin{theorem}
+\begin{thm}
 \label{5:theorem:CLJPc-neighbor-independence}
@@
 guarantees $w_i \ne w_j$.
-\end{theorem}
+\end{thm}
 \begin{proof}
@@
 Theorem~\ref{5:theorem:CLJPc-neighbor-independence}.
-\begin{corollary}
+\begin{cor}
 Any set of vertices in the graph of $S$ that share the same weight is
 an independent set.
-\end{corollary}
-\begin{corollary}
+\end{cor}
+\begin{cor}
 \label{5:cor:max-weight-indep-set}
@@
 neighborhood.
-\end{corollary}
+\end{cor}
 The first corollary states that independent sets can be selected in
@@
 Algorithm~\ref{5:alg:search}.
-\begin{algorithm}[h]
-\caption{Coarse-Grid Selection Graph Search}
-\label{5:alg:search}
-\textsc{Search-Graph}$(S,\, C,\, F)$ \{
-\begin{algorithmic}[1]
-\STATE $D \leftarrow \emptyset$
-\FORALL{$i \notin (C \cup F)$}
-\IF{$w_i > w_j,\,\forall j \in \mathcal{N}_i$}
-\STATE $D \leftarrow D \cup \{i\}$
-\ENDIF
-\ENDFOR
-\RETURN $D$
-\end{algorithmic}
-\}
-\end{algorithm}
+\begin{lstlisting}[caption={Coarse-Grid Selection Graph Search},label=5:alg:search]
+%\textsc{Search-Graph}$(S,\, C,\, F)$%
+$D \leftarrow \emptyset$
+for $i$ in $(C \cup F)$
+  if $w_i > w_j,\,\forall j \in \mathcal{N}_i$
+    $D \leftarrow D \cup \{i\}$
+  end
+end
+return $D$
+\end{lstlisting}
+% \begin{algorithm}[h]
+% \caption{Coarse-Grid Selection Graph Search}
+% \label{5:alg:search}
+% \textsc{Search-Graph}$(S,\, C,\, F)$ \{
+% \begin{algorithmic}[1]
+% \STATE $D \leftarrow \emptyset$
+% \FORALL{$i \notin (C \cup F)$}
+% \IF{$w_i > w_j,\,\forall j \in \mathcal{N}_i$}
+% \STATE $D \leftarrow D \cup \{i\}$
+% \ENDIF
+% \ENDFOR
+% \RETURN $D$
+% \end{algorithmic}
+% \}
+% \end{algorithm}
 In the first iteration, Line~3 is run $2|E|$ times. The total cost of
@@
 Pseudo-code for the weight update in CLJP is shown in Algorithm~\ref{5:alg:setup}.
-\begin{algorithm}[h]
-\caption{CLJP Weight Update for CSR Matrix}
-\label{5:alg:setup}
-\textsc{Update-Weights}$(S,\, D,\, C,\, F,\, w)$ \{
-\begin{algorithmic}[1]
-\FORALL{$d \in D$}
-\FORALL{$i \in S_d$}
-\STATE $w_i \leftarrow w_i - 1$ \COMMENT{see
-  Figure~\ref{3:fig:CLJP-update-heuristics} left}
-\STATE $S_d = S_d \setminus \{i\}$ \COMMENT{removing edge from graph}
-\ENDFOR
-\ENDFOR
-\FORALL{$i \notin (C \cup F)$}
-\FORALL{$k$ originally in $S_i$}
-\IF{$k \in D$}
-\STATE mark $k$ NEW-C-POINT
-\ENDIF
-\ENDFOR
-\FORALL{$j \in S_i$}
-\IF{$j \notin D$}
-\FORALL{$k \in S_j$}
-\IF[$i$ and $j$ mutually influenced by $k$]{$k$ is marked NEW-C-POINT}
-\STATE $w_j \leftarrow w_j - 1$ \COMMENT{see
-  Figure~\ref{3:fig:CLJP-update-heuristics} right}
-\STATE $S_i = S_i \setminus \{j\}$ \COMMENT{remove edge from $j$ to $i$}
-\ENDIF
-\ENDFOR
-\ENDIF
-\ENDFOR
-\FORALL{$k$ originally in $S_i$}
-\IF{$k \in D$}
-\STATE unmark $k$
-\STATE $S_i = S_i \setminus \{k\}$ \COMMENT{remove edge from $k$ to $i$,
-  if present}
-\ENDIF
-\ENDFOR
-\ENDFOR
-\end{algorithmic}
-\}
-\end{algorithm}
+\begin{lstlisting}[caption={CLJP Weight Update for CSR Matrix},label=5:alg:setup]
+%\textsc{Update-Weights}$(S,\, D,\, C,\, F,\, w)$%
+for $d \in D$
+  for $i \in S_d$
+    $w_i \leftarrow w_i - 1$    //see %Figure~\ref{3:fig:CLJP-update-heuristics} left%
+    $S_d = S_d \setminus \{i\}$ //removing edge from graph
+  end
+end
+for $i \notin (C \cup F)$
+  for $k$ originally in $S_i$
+    if $k \in D$
+      mark $k$ NEW-C-POINT
+    end
+  end
+  for $j \in S_i$
+    if $j \notin D$
+      for $k \in S_j$
+        if $k$ is marked NEW-C-POINT      // $i$ and $j$ mutually influenced by $k$
+          $w_j \leftarrow w_j - 1$        // see %Figure~\ref{3:fig:CLJP-update-heuristics} right%
+          $S_i = S_i \setminus \{j\}$     // remove edge from $j$ to $i$
+        end
+      end
+    end
+  end
+  for $k$ originally in $S_i$
+    if $k \in D$
+      unmark $k$
+      $S_i = S_i \setminus \{k\}$         // remove edge from $k$ to $i$, if present
+    end
+  end
+end
+\end{lstlisting}
+%\begin{algorithm}[h]
+%\caption{CLJP Weight Update for CSR Matrix}
+%\label{5:alg:setup}
+%\textsc{Update-Weights}$(S,\, D,\, C,\, F,\, w)$ \{
+%\begin{algorithmic}[1]
+%\FORALL{$d \in D$}
+%\FORALL{$i \in S_d$}
+%\STATE $w_i \leftarrow w_i - 1$ \COMMENT{see
+%  Figure~\ref{3:fig:CLJP-update-heuristics} left}
+%\STATE $S_d = S_d \setminus \{i\}$ \COMMENT{removing edge from graph}
+%\ENDFOR
+%\ENDFOR
+%\FORALL{$i \notin (C \cup F)$}
+%\FORALL{$k$ originally in $S_i$}
+%\IF{$k \in D$}
+%\STATE mark $k$ NEW-C-POINT
+%\ENDIF
+%\ENDFOR
+%\FORALL{$j \in S_i$}
+%\IF{$j \notin D$}
+%\FORALL{$k \in S_j$}
+%\IF[$i$ and $j$ mutually influenced by $k$]{$k$ is marked NEW-C-POINT}
+%\STATE $w_j \leftarrow w_j - 1$ \COMMENT{see
+%  Figure~\ref{3:fig:CLJP-update-heuristics} right}
+%\STATE $S_i = S_i \setminus \{j\}$ \COMMENT{remove edge from $j$ to $i$}
+%\ENDIF
+%\ENDFOR
+%\ENDIF
+%\ENDFOR
+%\FORALL{$k$ originally in $S_i$}
+%\IF{$k \in D$}
+%\STATE unmark $k$
+%\STATE $S_i = S_i \setminus \{k\}$ \COMMENT{remove edge from $k$ to $i$,
+%  if present}
+%\ENDIF
+%\ENDFOR
+%\ENDFOR
+%\end{algorithmic}
+%\}
+%\end{algorithm}
 The level of complication in this update routine is due to the CSR
@@
 neighborhoods}, are conceivable.
-\begin{definition}
+\begin{defn}
 The \emph{distance-$d$ neighborhood} of $i$, denoted
@@
 $\mathcal{N}_i^0 = \emptyset$.
-\end{definition}
+\end{defn}
 
@@
 selection neighborhood}.
-\begin{definition}
+\begin{defn}
 The \emph{extended selection neighborhood} of $i$, denoted
@@
 \mathcal{N}_i^s\right] \setminus \{i\}$.
-\end{definition}
+\end{defn}
 
@@
 assignments affect $w_i$. This is formalized below.
-\begin{lemma}
+\begin{lem}
 \label{5:lemma:C-lock}
@@
 the graph.
-\end{lemma}
+\end{lem}
 \begin{proof}
@@
 the first set of vertices satisfying the general selection conditions.
-\begin{definition}
+\begin{defn}
 Let $D_0$ be the set of all vertices satisfying
@@
 selection.
-\end{definition}
+\end{defn}
 
@@
 lemma.
-\begin{lemma}
+\begin{lem}
 \label{5:lemma:D0_invariant}
@@
 algorithms select vertices in $D_0$ as $C$-points.
-\end{lemma}
+\end{lem}
 \begin{proof}
@@
 
-\begin{theorem}
+\begin{thm}
 \label{5:theorem:invariance}
@@
 grids.
-\end{theorem}
+\end{thm}
 \begin{proof}
@@
 
-\begin{remark}
+\begin{rem}
 CLJP and CLJP-c use the distance-one neighborhood as the selection
@@
 CLJP-c select identical coarse grids.
-\end{remark}
+\end{rem}
 
@@
 outline the operations discussed above.
-\begin{algorithm}[!h]
-\caption{BSIS Data Structure Setup}
-\label{5:alg:BSIS-setup}
-\textsc{BSIS-Setup}$(S)$ \{
-\begin{algorithmic}[1]
-\FORALL{$i \in V$}
-\STATE $bucketID \leftarrow (w_i-1)\cdot numColors + color_i$
-\STATE $bucket[bucketID].\textsc{Insert}(i)$
-\ENDFOR
-\end{algorithmic}
-\}
-\end{algorithm}
-\begin{algorithm}[!h]
-\caption{Independent Set Selection}
-\label{5:alg:BSIS-indep-set-select}
-\textsc{BSIS-Independent-Set-Selection}$(S)$ \{
-\begin{algorithmic}[1]
-\RETURN non-empty $bucket$ with largest $bucketID$
-\end{algorithmic}
-\}
-\end{algorithm}
-\begin{algorithm}[!h]
-\caption{BSIS Weight Update}
-\label{5:alg:BSIS-weight-update}
-\textsc{BSIS-Weight-Update}$(S)$ \{
-\begin{algorithmic}[1]
-\STATE $bucketID \leftarrow (w_i-1)\cdot numColors + color_i$
-\STATE $bucket[bucketID].\textsc{Remove}(i)$
-\STATE $bucket[bucketID - numColors].\textsc{Insert}(i)$
-\end{algorithmic}
-\}
-\end{algorithm}
+\begin{lstlisting}[caption={BSIS Data Structure Setup},label=5:alg:BSIS-setup]
+%\textsc{BSIS-Setup}$(S)$%
+for $i \in V$
+  $bucketID \leftarrow (w_i-1)\cdot numColors + color_i$
+  $bucket[bucketID].\textsc{Insert}(i)$
+end
+\end{lstlisting}
+% \begin{algorithm}[!h]
+% \caption{BSIS Data Structure Setup}
+% \label{5:alg:BSIS-setup}
+% \textsc{BSIS-Setup}$(S)$ \{
+% \begin{algorithmic}[1]
+% \FORALL{$i \in V$}
+% \STATE $bucketID \leftarrow (w_i-1)\cdot numColors + color_i$
+% \STATE $bucket[bucketID].\textsc{Insert}(i)$
+% \ENDFOR
+% \end{algorithmic}
+% \}
+% \end{algorithm}
+\begin{lstlisting}[caption={Independent Set Selection},label=5:alg:BSIS-indep-set-select]
+%\textsc{BSIS-Independent-Set-Selection}$(S)$%
+return non-empty $bucket$ with largest $bucketID$
+\end{lstlisting}
+% \begin{algorithm}[!h]
+% \caption{Independent Set Selection}
+% \label{5:alg:BSIS-indep-set-select}
+% \textsc{BSIS-Independent-Set-Selection}$(S)$ \{
+% \begin{algorithmic}[1]
+% \RETURN non-empty $bucket$ with largest $bucketID$
+% \end{algorithmic}
+% \}
+% \end{algorithm}
+\begin{lstlisting}[caption={BSIS Weight Update},label=5:alg:BSIS-weight-update]
+%\textsc{BSIS-Weight-Update}$(S)$%
+$bucketID \leftarrow (w_i-1)\cdot numColors + color_i$
+$bucket[bucketID].\textsc{Remove}(i)$
+$bucket[bucketID - numColors].\textsc{Insert}(i)$
+\end{lstlisting}
+%\begin{algorithm}[!h]
+%\caption{BSIS Weight Update}
+%\label{5:alg:BSIS-weight-update}
+%\textsc{BSIS-Weight-Update}$(S)$ \{
+%\begin{algorithmic}[1]
+%\STATE $bucketID \leftarrow (w_i-1)\cdot numColors + color_i$
+%\STATE $bucket[bucketID].\textsc{Remove}(i)$
+%\STATE $bucket[bucketID - numColors].\textsc{Insert}(i)$
+%\end{algorithmic}
+%\}
+%\end{algorithm}
 \begin{figure}
   \begin{center}
-    %\includegraphics[width=0.8\textwidth]{images/5/bucket-algorithm.eps}
-    \includegraphics[width=0.98\textwidth]{images/5/BSIS/initial.eps}
+    %\includegraphics[width=0.8\textwidth]{images/bucket-algorithm}
+    \includegraphics[width=0.98\textwidth]{images/initial}
     \caption{The BSIS data structure.}
@@
   \begin{center}
-    \includegraphics[width=0.98\textwidth]{images/5/BSIS/cp1.eps}
+    \includegraphics[width=0.98\textwidth]{images/cp1}
     \caption{The BSIS data structure after selecting the first
@@
   \begin{center}
-    \includegraphics[width=0.98\textwidth]{images/5/BSIS/cp1-agg.eps}
+    \includegraphics[width=0.98\textwidth]{images/cp1-agg}
     \caption{The BSIS data structure after selecting the first
@@
   \begin{center}
-    \includegraphics[width=0.75\textwidth]{images/5/agg-no-scaling.eps}
+    \includegraphics[width=0.75\textwidth]{images/agg-no-scaling}
     \caption{Coarse-grid selection times using BSIS with aggregate
@@
   \begin{center}
-    \includegraphics[width=0.98\textwidth]{images/5/BSIS/cp1p.eps}
+    \includegraphics[width=0.98\textwidth]{images/cp1p}
     \caption{The painted boundary method with aggregate weight
@@
   \begin{center}
-    \includegraphics[width=0.98\textwidth]{images/5/BSIS/cp2p.eps}
+    \includegraphics[width=0.98\textwidth]{images/cp2p}
     \caption{The painted boundary method with aggregate weight
@@
 deCoughnyLB, SchloegelLB, ZoltanParHyp06ipdps, ZoltanParHypRepart07}.
+
+\end{document}

### 2007-08-06 bsis_paper/paper.tex
* Added image files.
git-svn-id: file:///home/rblake/svn/pamg/trunk@13 c7f3fe13-8c30-0410-9127-9a9c42fe009c
* This document class is strange and seems to be causing various   problems. Not sure what caused Luke to use this style. Need to   ask. * Cleaned up some of the problems with the references. * Cleaned up title page. * Thought about the design of the document and put some of that in   the TODO file. Began making changes in document structure in the   .tex file.
git-svn-id: file:///home/rblake/svn/pamg/trunk@14 c7f3fe13-8c30-0410-9127-9a9c42fe009c

@@
 % or use the graphicx package for more complicated commands
-\usepackage[draft]{graphicx}
+\usepackage{graphicx}
 % or use the epsfig package if you prefer to use the old commands
@@
 \title{Bucket-Sorted Independent Sets for Algebraic Multigrid}
-\author{David Alber\corauthref{cor}},
-\corauth[cor]{Corresponding author. Address: 201 N. Goodwin Ave., Siebel
-Center for Computer Science, Urbana, IL 61801}
-\ead{dalber@uiuc.edu}
-\ead[url]{http://www.cs.uiuc.edu/homes/dalber}
-\author{Luke Olson}
-\address{Department of Computer Science\\ University of Illinois at Urbana-Champaign}
-\ead{dalber@uiuc.edu}
+\author[NREL]{David Alber\corauthref{cor}},
+\corauth[cor]{Corresponding author. Address: 1617 Cole Blvd., Mail
+  Stop: 1608, Golden, CO 80401}
+\ead{david\_alber@nrel.gov}
+%\ead[url]{http://www.cs.uiuc.edu/homes/dalber}
+\author[UIUC]{Luke Olson}
+\address[NREL]{Scientific Computing Center\\ National Renewable Energy Laboratory}
+\address[UIUC]{Department of Computer Science\\ University of Illinois at Urbana-Champaign}
+\ead{lukeo@uiuc.edu}
 \ead[url]{http://www.cs.uiuc.edu/homes/lukeo}
@@
 presented.
+
+\section{Coarse Grid Selection}
+\label{sec:coarse-grid-select}
+
+\subsection{Problem}
+\label{sec:problem}
+% Outline coarse grid selection. Problem outline.
+
+\subsection{Independent Set-Based Selection}
+\label{sec:indep-set-based}
+% Talk about independent set-based algorithms here, and give general
+% independent set-based algrotihm here.
 
@@
     %\includegraphics[width=0.8\textwidth]{images/bucket-algorithm}
-    \includegraphics[width=0.98\textwidth]{images/initial}
+    \includegraphics[width=0.98\textwidth]{images/BSIS/initial}
     \caption{The BSIS data structure.}
@@
   \begin{center}
-    \includegraphics[width=0.98\textwidth]{images/cp1}
+    \includegraphics[width=0.98\textwidth]{images/BSIS/cp1}
     \caption{The BSIS data structure after selecting the first
@@
   \begin{center}
-    \includegraphics[width=0.98\textwidth]{images/cp1-agg}
+    \includegraphics[width=0.98\textwidth]{images/BSIS/cp1-agg}
     \caption{The BSIS data structure after selecting the first
@@
 information on solve phase performance for AMG with CLJP-c is
-documented in~\cite{alber-cljpc, alber-PCGS}.
+documented in~\cite{alber-cljpc,alber-PCGS}.
 
@@
   \begin{center}
-    \includegraphics[width=0.98\textwidth]{images/cp1p}
+    \includegraphics[width=0.98\textwidth]{images/BSIS/cp1p}
     \caption{The painted boundary method with aggregate weight
@@
   \begin{center}
-    \includegraphics[width=0.98\textwidth]{images/cp2p}
+    \includegraphics[width=0.98\textwidth]{images/BSIS/cp2p}
     \caption{The painted boundary method with aggregate weight
@@
 and third ideas are similar in nature and both involve using dynamic
-load balancing techniques~\cite{DevineBomanKarypisPP04, CybDLB,
-deCoughnyLB, SchloegelLB, ZoltanParHyp06ipdps, ZoltanParHypRepart07}.
+load balancing
+techniques~\cite{DevineBomanKarypisPP04,CybDLB,deCoughnyLB,SchloegelLB,ZoltanParHyp06ipdps,ZoltanParHypRepart07}.
+
+\bibliographystyle{elsart-num}
+\bibliography{paper}
 

### 2007-08-12 bsis_paper/paper.tex
* Rearranged sections. * Writing work on first two sections. * Cleaned up .bib file.
git-svn-id: file:///home/rblake/svn/pamg/trunk@15 c7f3fe13-8c30-0410-9127-9a9c42fe009c

@@
 
+\newcommand{\hp}{H1$^\prime$}
+
 \begin{document}
@@
 \ead{david\_alber@nrel.gov}
-%\ead[url]{http://www.cs.uiuc.edu/homes/dalber}
+\ead[url]{http://google.com/ie?q=david+alber+nrel\&num=1}
 \author[UIUC]{Luke Olson}
@@
 \begin{abstract}
-% Text of abstract
 \end{abstract}
@@
 
-% main text
+\section{Introduction}
+\label{sec:intro}
+The algebraic multigrid (AMG) method~\cite{Ruge1987,Brandt1982} is an
+efficient numerical algorithm for finding the solution to numerous
+problems, including linear systems $Ax = b$ arising from the
+discretization of partial differential equations on structured and
+unstructured meshes. In many cases, the computational complexity of
+AMG is $O(n)$, where $n$ is the number of unknowns in the linear
+system. The linear cost scaling property of AMG makes the method
+attractive, especially for large problems.
 
-\section{Introduction}
-Chapters~\ref{chap:CGS} and~\ref{chap:color-based} demonstrate the
-behavior of independent set-based coarse-grid selection
-algorithms. One common design property among the selection algorithms
-is a routine to search for vertices to become
-$C$-points. Additionally, an update in the weights of vertices as the
-coarse-grid selection often follows. In this chapter, new theory and
-algorithms to decrease the computational cost associated with the
-search and weight update procedures used in coarse-grid selection are
-presented.
+AMG algorithms are generally divided into two phases: the setup phase
+and the solve phase. Setup phase algorithms provide functionality for
+constructing elements needed by the solve phase, such as coarse grids,
+operators to transfer information between levels of the problem, and
+the linear systems to be solved on each level. The solve phase applies
+relaxation methods on the problems generated in the setup phase and
+uses the transfer operators to move vectors between levels.
 
-\section{Coarse Grid Selection}
-\label{sec:coarse-grid-select}
+In this paper, we focus attention on coarse-grid selection algorithms
+in the AMG setup phase. One class of coarse-grid selection algorithms,
+independent set-based algorithms, is studied in detail. In
+Section~\ref{sec:cgs}, we introduce the coarse-grid selection problems
+and the goals of coarse-grid selection algorithms. The components of
+independent set-based coarsening algorithms are discussed.
 
-\subsection{Problem}
-\label{sec:problem}
-% Outline coarse grid selection. Problem outline.
+\section{Coarse-Grid Selection}
+\label{sec:cgs}
+In AMG, coarse-grid selection is the process of selecting the degrees
+of freedom for the coarse level problem. In classical AMG, degrees of
+freedom on a coarse level are a subset of the degrees of freedom on a
+finer level. Smoothed
+aggregation~\cite{Vanek1996,Vanek2001,brezina2005} forms a
+coarse-level degree of freedom by aggregating several fine-level
+degrees of freedom. In this paper, we study coarsening for classical
+AMG.
+
+Given a set of fine-level degrees of freedom, $\Omega_k$, coarse-grid
+selection for classical AMG selects a set of coarse-level degrees of
+freedom $\Omega_{k+1} \subset \Omega_k$. Several approaches have been
+developed for selecting the coarse grid.
+
+Independent set-based algorithms are most numerous and all trace their
+roots to the classical AMG coarsening algorithm, RS
+coarsening~\cite{Ruge1987}. These methods are our focus and are
+discussed in more detail in the next section.
+
+Other methods take substantially different approaches to
+coarsening. The coarsening technique in~\cite{Griebel2006} develops a
+parallel coarsening algorithm that utilizes RS. On each processor, RS
+selects several different coarse grids, and then one coarse grid is
+selected per processor in a way to minimize special treatment on
+processor boundaries. In~\cite{MacLachlan2006}, a greedy approach is
+used to produce a good splitting of the unknowns into fine-grid and
+coarse-grid sets. Subdomain blocking techniques~\cite{Krechel2001}
+offer another approach for parallel coarse-grid selection by
+decoupling coarse grids and alleviating the need for communication on
+coarse levels. Compatible relaxation
+approaches~\cite{Brandt2000,Livne2004,Brannick2005} form coarse grids
+incrementally and decide when the coarse grid is sufficient through a
+quality measure based on relaxation.
 
@@
 \label{sec:indep-set-based}
-% Talk about independent set-based algorithms here, and give general
-% independent set-based algrotihm here.
+Independent set-based coarse grid selection algorithms rely on
+\emph{strength of connection} measures to ascertain the influence
+between unknowns.
 
-\section{CLJP-c}
-Recall that CLJP-c colors the graph of $S$ before selecting
-$C$-points, and the colors are used as one component of the vertex
-weights. As a result, the structure of the coarse grids selected is
-improved. More formally, the use of graph coloring provides the
-following important result.
-\begin{thm}
-\label{5:theorem:CLJPc-neighbor-independence}
-For all pairs of vertices $i$ and $j \in \mathcal{N}_i$ CLJP-c
-guarantees $w_i \ne w_j$.
-\end{thm}
-\begin{proof}
-Assume two adjacent vertices $i$ and $j$ have the same weight. That is,
-$|S_i^T| = |S_j^T|$ and the weight augmentation provided through
-coloring is the same for $i$ and $j$. The graph of $S$, however, is
-colored such that $i$ and $j$ are different colors for all $j \in
-\mathcal{N}_i$, so a contradiction is reached.
-\end{proof}
-Theorem~\ref{5:theorem:CLJPc-neighbor-independence} establishes that
-all adjacent vertices have different weights in CLJP-c, which is not
-guaranteed in CLJP, although is unlikely to occur. The following
-corollaries are a result of
-Theorem~\ref{5:theorem:CLJPc-neighbor-independence}.
-\begin{cor}
-Any set of vertices in the graph of $S$ that share the same weight is
-an independent set.
-\end{cor}
-\begin{cor}
-\label{5:cor:max-weight-indep-set}
-The set of vertices with the largest weight in the graph of $S$ form
-an independent set satisfying~\ref{5:eqn:D-conditions}. That is, each
-vertex in that independent set has a uniquely maximal weight in its
-neighborhood.
-\end{cor}
-The first corollary states that independent sets can be selected in
-the graph simply by selecting sets of vertices with the same
-weight. Corollary~\ref{5:cor:max-weight-indep-set} refines this
-observation to a subset of vertices guaranteed to satisfy the
-selection criterion. In particular, it shows it is possible to build
-the coarse grid by selecting vertices with the maximum weight,
-updating weights, selecting the next set of vertices with maximum
-weight, and so on. This approach is taken by the algorithm developed
-in Section~\ref{5:sec:contribution}. It is proven in
-Section~\ref{5:sec:cg-invariance} that despite the difference in
-approach, the resulting coarse grid is the same as that selected by
-CLJP-c.
+Many independent set-based coarsening algorithms form $\Omega_{k+1}$
+using the following heuristics.
+\begin{description}
+\item[H1:] For each unknown $j$ that strongly influences $F$-point
+  $i$, $j$ is either a $C$-point or strongly depends on a $C$-point
+  $k$ that also strongly influences $i$.
+\item[H2:] The set of $C$-points needs to form a maximal independent
+  set in the reduced graph of $S$ such that no $C$-point strongly
+  depends on another $C$-point.
+\end{description}
+H1 and H2 cannot, in general, be satisfied simultaneously, so H1 is
+enforced by independent set-based methods, while H2 serves as a
+guideline whose purpose is to encourage selection of small
+$|\Omega_{k+1}|$.
 
-\section{Coarse-Grid Selection Search and Weight Update}
-\label{5:sec:s-and-w-update}
-CLJP and its descendents select an independent set $D$ in each
-iteration. Stated originally in Section~\ref{3:sec:CLJP}, the
-condition for selecting $D$ is that all vertices $i \in D$ must
-satisfy
-\begin{equation}
-\label{5:eqn:D-conditions}
-w_i > w_j\;\textrm{for all } j \in \mathcal{N}_i.
-\end{equation}
-CLJP and CLJP-c rely on a search routine to locate locally maximal
-weights and on a weight update routine to modify weights of vertices
-connected to new $C$-points.
+Some independent set-based coarsening algorithms use a variant of H1,
+called \hp, which is a relaxed version of H1.
+\begin{description}
+\item[\hp] Each $F$-point must be strongly influenced by at least one
+  $C$-point.
+\end{description}
 
-The algorithms for searching and updating vertex weights in CLJP in
-detail are examined in this section. In particular, the impact of
-using a sparse matrix format on the coarsening procedure. The
-pseudo-code below assumes the software uses a compressed sparse row
-(CSR)~\cite{saadBook} matrix format or other similar format, which are
-common matrix formats in numerical software. CSR provides low memory
-costs for storing sparse matrices and provides efficient access to the
-nonzeros in a row. Accessing the nonzeros in a column is an expensive
-operation in this format and strongly influences the weight update
-routine in CLJP-style algorithms because $S$ is, in general, not
-symmetric.
+\section{Invariance of Selection in Independent Set-Based Methods}
+\label{sec:invariance}
 
-The search step in CLJP is implemented as shown in
-Algorithm~\ref{5:alg:search}.
-\begin{lstlisting}[caption={Coarse-Grid Selection Graph Search},label=5:alg:search]
-%\textsc{Search-Graph}$(S,\, C,\, F)$%
-$D \leftarrow \emptyset$
-for $i$ in $(C \cup F)$
-  if $w_i > w_j,\,\forall j \in \mathcal{N}_i$
-    $D \leftarrow D \cup \{i\}$
-  end
-end
-return $D$
-\end{lstlisting}
-% \begin{algorithm}[h]
-% \caption{Coarse-Grid Selection Graph Search}
-% \label{5:alg:search}
-% \textsc{Search-Graph}$(S,\, C,\, F)$ \{
-% \begin{algorithmic}[1]
-% \STATE $D \leftarrow \emptyset$
-% \FORALL{$i \notin (C \cup F)$}
-% \IF{$w_i > w_j,\,\forall j \in \mathcal{N}_i$}
-% \STATE $D \leftarrow D \cup \{i\}$
-% \ENDIF
-% \ENDFOR
-% \RETURN $D$
-% \end{algorithmic}
-% \}
-% \end{algorithm}
-In the first iteration, Line~3 is run $2|E|$ times. The total cost of
-search in constructing the coarse grid depends on the number of
-iterations needed. Even in the best case of $\Omega(E)$ time, the cost
-is significant when the graph contains large numbers of edges, as
-usually happens on the lower levels in the grid hierarchy
-(see~\cite{alber-PCGS} for examples). In the next section, a new
-technique is introduced for conducting the search in coarse-grid
-selection algorithms independent of the number of edges in the graph.
-
-Pseudo-code for the weight update in CLJP is shown in Algorithm~\ref{5:alg:setup}.
-\begin{lstlisting}[caption={CLJP Weight Update for CSR Matrix},label=5:alg:setup]
-%\textsc{Update-Weights}$(S,\, D,\, C,\, F,\, w)$%
-for $d \in D$
-  for $i \in S_d$
-    $w_i \leftarrow w_i - 1$    //see %Figure~\ref{3:fig:CLJP-update-heuristics} left%
-    $S_d = S_d \setminus \{i\}$ //removing edge from graph
-  end
-end
-for $i \notin (C \cup F)$
-  for $k$ originally in $S_i$
-    if $k \in D$
-      mark $k$ NEW-C-POINT
-    end
-  end
-  for $j \in S_i$
-    if $j \notin D$
-      for $k \in S_j$
-        if $k$ is marked NEW-C-POINT      // $i$ and $j$ mutually influenced by $k$
-          $w_j \leftarrow w_j - 1$        // see %Figure~\ref{3:fig:CLJP-update-heuristics} right%
-          $S_i = S_i \setminus \{j\}$     // remove edge from $j$ to $i$
-        end
-      end
-    end
-  end
-  for $k$ originally in $S_i$
-    if $k \in D$
-      unmark $k$
-      $S_i = S_i \setminus \{k\}$         // remove edge from $k$ to $i$, if present
-    end
-  end
-end
-\end{lstlisting}
-%\begin{algorithm}[h]
-%\caption{CLJP Weight Update for CSR Matrix}
-%\label{5:alg:setup}
-%\textsc{Update-Weights}$(S,\, D,\, C,\, F,\, w)$ \{
-%\begin{algorithmic}[1]
-%\FORALL{$d \in D$}
-%\FORALL{$i \in S_d$}
-%\STATE $w_i \leftarrow w_i - 1$ \COMMENT{see
-%  Figure~\ref{3:fig:CLJP-update-heuristics} left}
-%\STATE $S_d = S_d \setminus \{i\}$ \COMMENT{removing edge from graph}
-%\ENDFOR
-%\ENDFOR
-%\FORALL{$i \notin (C \cup F)$}
-%\FORALL{$k$ originally in $S_i$}
-%\IF{$k \in D$}
-%\STATE mark $k$ NEW-C-POINT
-%\ENDIF
-%\ENDFOR
-%\FORALL{$j \in S_i$}
-%\IF{$j \notin D$}
-%\FORALL{$k \in S_j$}
-%\IF[$i$ and $j$ mutually influenced by $k$]{$k$ is marked NEW-C-POINT}
-%\STATE $w_j \leftarrow w_j - 1$ \COMMENT{see
-%  Figure~\ref{3:fig:CLJP-update-heuristics} right}
-%\STATE $S_i = S_i \setminus \{j\}$ \COMMENT{remove edge from $j$ to $i$}
-%\ENDIF
-%\ENDFOR
-%\ENDIF
-%\ENDFOR
-%\FORALL{$k$ originally in $S_i$}
-%\IF{$k \in D$}
-%\STATE unmark $k$
-%\STATE $S_i = S_i \setminus \{k\}$ \COMMENT{remove edge from $k$ to $i$,
-%  if present}
-%\ENDIF
-%\ENDFOR
-%\ENDFOR
-%\end{algorithmic}
-%\}
-%\end{algorithm}
-The level of complication in this update routine is due to the CSR
-format and the need to find vertices strongly influenced by new
-$C$-points. When a new $C$-point $k$ is selected, the first type of
-weight update is trivial since determining the vertices in $S_k$ is
-inexpensive. The second type of update is more expensive since the
-vertices influenced by $k$ are difficult to determine in a CSR
-format. The update requires a search of many vertices $i$ and all of
-their strong influencing neighbors $j$. The routine then searches
-strongly influencing $j$ to determine if any $k \in D$ strongly
-influences both $i$ and $j$. The cost increases dramatically as the
-density of $S$ increases. Large operator complexities have a
-disproportionately large impact on coarse-grid selection run time. In
-the next section, a modified update routine to compliment the new
-search technique is introduced.
-
-\section{Bucket Sorted Independent Set Selection}
-\label{5:sec:contribution}
-In this section, new techniques for searching the graph of $S$ for new
-$C$-points and subsequently updating the weights of remaining vertices
-are developed. The new algorithm is labeled Bucket Sorted Independent
-Sets (BSIS) to reflect the data structure used.
-
-Like CLJP-c, BSIS depends on graph coloring, but utilizes modified
-routines for search and weight update. Furthermore, rather than
-applying the color information to augment vertex weights, BSIS uses
-the colors in a bucket data structure. Once initialized, this data
-structure selects independent sets, which satisfy the conditions in
-(\ref{5:eqn:D-conditions}), in constant time.
-
-\subsection{Coarse Grid Invariance}
-\label{5:sec:cg-invariance}
 Theory is developed in this section to prove coarse-grid selection
@@
 
+\section{Bucket Sorted Independent Set Selection}
+\label{5:sec:contribution}
+
+\subsection{CLJP-c}
+Recall that CLJP-c colors the graph of $S$ before selecting
+$C$-points, and the colors are used as one component of the vertex
+weights. As a result, the structure of the coarse grids selected is
+improved. More formally, the use of graph coloring provides the
+following important result.
+\begin{thm}
+\label{5:theorem:CLJPc-neighbor-independence}
+For all pairs of vertices $i$ and $j \in \mathcal{N}_i$ CLJP-c
+guarantees $w_i \ne w_j$.
+\end{thm}
+\begin{proof}
+Assume two adjacent vertices $i$ and $j$ have the same weight. That is,
+$|S_i^T| = |S_j^T|$ and the weight augmentation provided through
+coloring is the same for $i$ and $j$. The graph of $S$, however, is
+colored such that $i$ and $j$ are different colors for all $j \in
+\mathcal{N}_i$, so a contradiction is reached.
+\end{proof}
+Theorem~\ref{5:theorem:CLJPc-neighbor-independence} establishes that
+all adjacent vertices have different weights in CLJP-c, which is not
+guaranteed in CLJP, although is unlikely to occur. The following
+corollaries are a result of
+Theorem~\ref{5:theorem:CLJPc-neighbor-independence}.
+\begin{cor}
+Any set of vertices in the graph of $S$ that share the same weight is
+an independent set.
+\end{cor}
+\begin{cor}
+\label{5:cor:max-weight-indep-set}
+The set of vertices with the largest weight in the graph of $S$ form
+an independent set satisfying~\ref{5:eqn:D-conditions}. That is, each
+vertex in that independent set has a uniquely maximal weight in its
+neighborhood.
+\end{cor}
+The first corollary states that independent sets can be selected in
+the graph simply by selecting sets of vertices with the same
+weight. Corollary~\ref{5:cor:max-weight-indep-set} refines this
+observation to a subset of vertices guaranteed to satisfy the
+selection criterion. In particular, it shows it is possible to build
+the coarse grid by selecting vertices with the maximum weight,
+updating weights, selecting the next set of vertices with maximum
+weight, and so on. This approach is taken by the algorithm developed
+in Section~\ref{5:sec:contribution}. It is proven in
+Section~\ref{5:sec:cg-invariance} that despite the difference in
+approach, the resulting coarse grid is the same as that selected by
+CLJP-c.
+
+\subsection{BSIS}
+\label{sec:bsis}
+In this section, new techniques for searching the graph of $S$ for new
+$C$-points and subsequently updating the weights of remaining vertices
+are developed. The new algorithm is labeled Bucket Sorted Independent
+Sets (BSIS) to reflect the data structure used.
+
+Like CLJP-c, BSIS depends on graph coloring, but utilizes modified
+routines for search and weight update. Furthermore, rather than
+applying the color information to augment vertex weights, BSIS uses
+the colors in a bucket data structure. Once initialized, this data
+structure selects independent sets, which satisfy the conditions in
+(\ref{5:eqn:D-conditions}), in constant time.
+
+\subsection{Motivation}
+\label{sec:motivation}
+CLJP and its descendents select an independent set $D$ in each
+iteration. Stated originally in Section~\ref{3:sec:CLJP}, the
+condition for selecting $D$ is that all vertices $i \in D$ must
+satisfy
+\begin{equation}
+\label{5:eqn:D-conditions}
+w_i > w_j\;\textrm{for all } j \in \mathcal{N}_i.
+\end{equation}
+CLJP and CLJP-c rely on a search routine to locate locally maximal
+weights and on a weight update routine to modify weights of vertices
+connected to new $C$-points.
+
+The algorithms for searching and updating vertex weights in CLJP in
+detail are examined in this section. In particular, the impact of
+using a sparse matrix format on the coarsening procedure. The
+pseudo-code below assumes the software uses a compressed sparse row
+(CSR)~\cite{saadBook} matrix format or other similar format, which are
+common matrix formats in numerical software. CSR provides low memory
+costs for storing sparse matrices and provides efficient access to the
+nonzeros in a row. Accessing the nonzeros in a column is an expensive
+operation in this format and strongly influences the weight update
+routine in CLJP-style algorithms because $S$ is, in general, not
+symmetric.
+
+The search step in CLJP is implemented as shown in
+Algorithm~\ref{5:alg:search}.
+\begin{lstlisting}[caption={Coarse-Grid Selection Graph Search},label=5:alg:search]
+%\textsc{Search-Graph}$(S,\, C,\, F)$%
+$D \leftarrow \emptyset$
+for $i$ in $(C \cup F)$
+  if $w_i > w_j,\,\forall j \in \mathcal{N}_i$
+    $D \leftarrow D \cup \{i\}$
+  end
+end
+return $D$
+\end{lstlisting}
+% \begin{algorithm}[h]
+% \caption{Coarse-Grid Selection Graph Search}
+% \label{5:alg:search}
+% \textsc{Search-Graph}$(S,\, C,\, F)$ \{
+% \begin{algorithmic}[1]
+% \STATE $D \leftarrow \emptyset$
+% \FORALL{$i \notin (C \cup F)$}
+% \IF{$w_i > w_j,\,\forall j \in \mathcal{N}_i$}
+% \STATE $D \leftarrow D \cup \{i\}$
+% \ENDIF
+% \ENDFOR
+% \RETURN $D$
+% \end{algorithmic}
+% \}
+% \end{algorithm}
+In the first iteration, Line~3 is run $2|E|$ times. The total cost of
+search in constructing the coarse grid depends on the number of
+iterations needed. Even in the best case of $\Omega(E)$ time, the cost
+is significant when the graph contains large numbers of edges, as
+usually happens on the lower levels in the grid hierarchy
+(see~\cite{alber-PCGS} for examples). In the next section, a new
+technique is introduced for conducting the search in coarse-grid
+selection algorithms independent of the number of edges in the graph.
+
+Pseudo-code for the weight update in CLJP is shown in Algorithm~\ref{5:alg:setup}.
+\begin{lstlisting}[caption={CLJP Weight Update for CSR Matrix},label=5:alg:setup]
+%\textsc{Update-Weights}$(S,\, D,\, C,\, F,\, w)$%
+for $d \in D$
+  for $i \in S_d$
+    $w_i \leftarrow w_i - 1$    //see %Figure~\ref{3:fig:CLJP-update-heuristics} left%
+    $S_d = S_d \setminus \{i\}$ //removing edge from graph
+  end
+end
+for $i \notin (C \cup F)$
+  for $k$ originally in $S_i$
+    if $k \in D$
+      mark $k$ NEW-C-POINT
+    end
+  end
+  for $j \in S_i$
+    if $j \notin D$
+      for $k \in S_j$
+        if $k$ is marked NEW-C-POINT      // $i$ and $j$ mutually influenced by $k$
+          $w_j \leftarrow w_j - 1$        // see %Figure~\ref{3:fig:CLJP-update-heuristics} right%
+          $S_i = S_i \setminus \{j\}$     // remove edge from $j$ to $i$
+        end
+      end
+    end
+  end
+  for $k$ originally in $S_i$
+    if $k \in D$
+      unmark $k$
+      $S_i = S_i \setminus \{k\}$         // remove edge from $k$ to $i$, if present
+    end
+  end
+end
+\end{lstlisting}
+%\begin{algorithm}[h]
+%\caption{CLJP Weight Update for CSR Matrix}
+%\label{5:alg:setup}
+%\textsc{Update-Weights}$(S,\, D,\, C,\, F,\, w)$ \{
+%\begin{algorithmic}[1]
+%\FORALL{$d \in D$}
+%\FORALL{$i \in S_d$}
+%\STATE $w_i \leftarrow w_i - 1$ \COMMENT{see
+%  Figure~\ref{3:fig:CLJP-update-heuristics} left}
+%\STATE $S_d = S_d \setminus \{i\}$ \COMMENT{removing edge from graph}
+%\ENDFOR
+%\ENDFOR
+%\FORALL{$i \notin (C \cup F)$}
+%\FORALL{$k$ originally in $S_i$}
+%\IF{$k \in D$}
+%\STATE mark $k$ NEW-C-POINT
+%\ENDIF
+%\ENDFOR
+%\FORALL{$j \in S_i$}
+%\IF{$j \notin D$}
+%\FORALL{$k \in S_j$}
+%\IF[$i$ and $j$ mutually influenced by $k$]{$k$ is marked NEW-C-POINT}
+%\STATE $w_j \leftarrow w_j - 1$ \COMMENT{see
+%  Figure~\ref{3:fig:CLJP-update-heuristics} right}
+%\STATE $S_i = S_i \setminus \{j\}$ \COMMENT{remove edge from $j$ to $i$}
+%\ENDIF
+%\ENDFOR
+%\ENDIF
+%\ENDFOR
+%\FORALL{$k$ originally in $S_i$}
+%\IF{$k \in D$}
+%\STATE unmark $k$
+%\STATE $S_i = S_i \setminus \{k\}$ \COMMENT{remove edge from $k$ to $i$,
+%  if present}
+%\ENDIF
+%\ENDFOR
+%\ENDFOR
+%\end{algorithmic}
+%\}
+%\end{algorithm}
+The level of complication in this update routine is due to the CSR
+format and the need to find vertices strongly influenced by new
+$C$-points. When a new $C$-point $k$ is selected, the first type of
+weight update is trivial since determining the vertices in $S_k$ is
+inexpensive. The second type of update is more expensive since the
+vertices influenced by $k$ are difficult to determine in a CSR
+format. The update requires a search of many vertices $i$ and all of
+their strong influencing neighbors $j$. The routine then searches
+strongly influencing $j$ to determine if any $k \in D$ strongly
+influences both $i$ and $j$. The cost increases dramatically as the
+density of $S$ increases. Large operator complexities have a
+disproportionately large impact on coarse-grid selection run time. In
+the next section, a modified update routine to compliment the new
+search technique is introduced.
+
 \subsection{Bucket Sorted Independent Sets Algorithm}
@@
 
-\section{Weight Update Aggregation}
+\subsection{Weight Update Aggregation}
 \label{5:sec:agg}
@@
 
-\section{Experimental Results}
+\subsection{Experimental Results}
 \label{5:sec:experiments}
@@
 
-\section{BSIS Variants}
+\subsection{BSIS Variants}
 The ideas developed in this chapter are applicable to other coarsening
@@
 
-\section{Parallelizing BSIS}
+\subsection{Parallelizing BSIS}
 Using BSIS in a parallel algorithm presents challenges because the
@@
 
+\section{Conclusions}
+\label{sec:conclusions}
+
 \bibliographystyle{elsart-num}

### 2007-08-21 petsc_amg/debug_std.py
Adding the configure script I used for petsc.  Configuring with this script will allow you to compare yoru results to hypre's AMG.
git-svn-id: file:///home/rblake/svn/pamg/trunk@16 c7f3fe13-8c30-0410-9127-9a9c42fe009c

@@ -1,0 +1,24 @@
+#!/usr/bin/env python
+
+if __name__ == '__main__':
+  import sys
+  import os
+  sys.path.insert(0, os.getcwd()+'/config')
+  import configure
+  PETSC_ARCH = 'debug_std'
+  del os.environ['PETSC_DIR']
+  del os.environ['PETSC_ARCH']
+  configure_options = ['--with-petsc-arch='+PETSC_ARCH,
+                       '--download-ml=yes',
+                       '-PETSC_ARCH='+PETSC_ARCH,
+                       '--download-hypre=yes',
+                       '--download-mpich=yes',
+                       ]
+  configure.petsc_configure(configure_options)
+
+
+#                       '--download-superlu_dist=yes',
+#                       '--download-superlu=yes',
+#                       '--download-scalapack=yes',
+#                       '--download-blacs=ifneeded',
+#                       '--download-mumps=yes',

### 2007-08-21 petsc_amg/luke.c
added a test to play around with shell contexts in PETsc
git-svn-id: file:///home/rblake/svn/pamg/trunk@17 c7f3fe13-8c30-0410-9127-9a9c42fe009c

@@ -1,0 +1,333 @@
+/* 
+ ------
+ log:
+  lukeo:  Tue Aug 21 10:31:15 CDT 2007
+      - touched
+ ------
+ 
+ use
+ mpirun -np 8 ./luke args
+ or
+ $PETSC_DIR/bin/petscmpirun -np 8 ./luke args
+ 
+ */
+#include "petscksp.h"
+
+static char help[] = "Easy MG\n\
+                      Input Parameters:\n\
+                        -kx : integer for sin wave exact soln in x-dir\n\
+                        -ky : integer for sin wave exact soln in y-dir\n\
+                        -nx : mesh points in x-dir\n\
+                        -ny : mesh points in y-dir\n\";
+
+/* TODO: move SCAMG stuff to scamg.h */
+
+/* context for MG preconditioner (0 level) */
+typedef struct {
+  Vec diag;
+} SCAMG;
+
+/* declarations for SCAMG routines */
+extern PetscErrorCode SCAMGCreate(SCAMG**);
+extern PetscErrorCode SCAMGSetUp(SCAMG*,Mat,Vec);
+extern PetscErrorCode SCAMGApply(void*,Vec x,Vec y);
+extern PetscErrorCode SCAMGDestroy(SCAMG*);
+
+#undef __FUNCT__  
+#define __FUNCT__ "main"
+int main(int argc,char **args)
+{
+  Vec            x,b,xex;   /* approx solution, RHS, exact solution */
+  Mat            A;         /* linear system matrix */
+  KSP            ksp;       /* linear solver context */
+  PC             pc;        /* preconditioner context */
+  PetscReal      norm;      /* norm of solution error */
+  SCAMG          *SCAMGsh;  /* MG context */
+  PetscScalar    v,one = 1.0,neg_one = -1.0;
+  PetscInt       i,j,Ii,J,Istart,Iend,nx = 10,ny = 8,its,kx=1,ky=1;
+  PetscErrorCode ierr;
+  PetscTruth     use_SCAMG;
+
+  PetscInitialize(&argc,&args,(char *)0,help);
+  ierr = PetscOptionsGetInt(PETSC_NULL,"-nx",&nx,PETSC_NULL);CHKERRQ(ierr);
+  ierr = PetscOptionsGetInt(PETSC_NULL,"-ny",&ny,PETSC_NULL);CHKERRQ(ierr);
+
[…]

### 2007-08-22 petsc_amg/luke.c
0 level MG now works :)   In fact there's an ex in tutorials just about like this one
git-svn-id: file:///home/rblake/svn/pamg/trunk@18 c7f3fe13-8c30-0410-9127-9a9c42fe009c

@@ -1,8 +1,11 @@
 /* 
  ------
- log:
-  lukeo:  Tue Aug 21 10:31:15 CDT 2007
+ log (use :r!date in vim):
+  lukeo: Wed Aug 22 09:15:07 CDT 2007
+      - started working on the shell context for AMG
+  lukeo: Tue Aug 21 10:31:15 CDT 2007
       - touched
+      - copied some structure from tutorial/ex*.c
  ------
  
  use
@@ -11,6 +14,7 @@
  $PETSC_DIR/bin/petscmpirun -np 8 ./luke args
  
  */
+#include "petscda.h"
 #include "petscksp.h"
 
 static char help[] = "Easy MG\n\
@@ -18,41 +22,56 @@ static char help[] = "Easy MG\n\
                         -kx : integer for sin wave exact soln in x-dir\n\
                         -ky : integer for sin wave exact soln in y-dir\n\
                         -nx : mesh points in x-dir\n\
-                        -ny : mesh points in y-dir\n\";
+                        -ny : mesh points in y-dir\n";
 
-/* TODO: move SCAMG stuff to scamg.h */
+/* TODO: move ScAMG stuff to scamg.h */
 
 /* context for MG preconditioner (0 level) */
 typedef struct {
-  Vec diag;
-} SCAMG;
+  Vec diaginv;
+} ScAMG;
 
-/* declarations for SCAMG routines */
-extern PetscErrorCode SCAMGCreate(SCAMG**);
-extern PetscErrorCode SCAMGSetUp(SCAMG*,Mat,Vec);
-extern PetscErrorCode SCAMGApply(void*,Vec x,Vec y);
-extern PetscErrorCode SCAMGDestroy(SCAMG*);
+/* declarations for ScAMG routines */
+extern PetscErrorCode ScAMGCreate(ScAMG**);
+extern PetscErrorCode ScAMGSetUp(ScAMG*,Mat,Vec);
+extern PetscErrorCode ScAMGApply(void*,Vec x,Vec y);
+extern PetscErrorCode ScAMGDestroy(ScAMG*);
 
 #undef __FUNCT__  
 #define __FUNCT__ "main"
-int main(int argc,char **args)
-{
[…]

### 2007-08-24 bsis_paper/paper.tex
* Section 2 updates and new content.
git-svn-id: file:///home/rblake/svn/pamg/trunk@19 c7f3fe13-8c30-0410-9127-9a9c42fe009c

@@
 \label{sec:indep-set-based}
-Independent set-based coarse grid selection algorithms rely on
-\emph{strength of connection} measures to ascertain the influence
-between unknowns.
+Independent set-based coarse grid selection algorithms implement three
+routines to select coarse grids: initialization, independent set
+selection, and weight update. A general independent-set based
+coarsening algorithm is shown in
+Algorithm~\ref{alg:general-indep-set}.
+\begin{lstlisting}[caption={Independent Set-Based Coarse Grid
+Selection},label=alg:general-indep-set]
+%\textsc{Indep-Set-Coarse-Grid-Selection}$(\Omega)$:%
+$F \leftarrow \emptyset$
+$C \leftarrow \emptyset$
+$w \leftarrow\,$%\textsc{Initialize-Weights}$(\Omega)$%
+while $\Omega \ne \emptyset$
+   $D \leftarrow\,$%\textsc{Select-Indep-Set}$(w,\, \Omega)$%
+   $(F,\, C,\, \Omega,\, w) \leftarrow\,$%\textsc{Update-Weights}$(F,\,
+                                C,\, \Omega,\, w)$%
+end
+\end{lstlisting}
+
+Each vertex in the graph is assigned a weight based on its merit to
+becoming a $C$-point. The largest weights are conferred to vertices
+best suited to become $C$-points. The quality of potential $C$-points
+is quantified by \emph{strength of connection} measures. The classical
+strength of connection measure is based entirely on the relative sizes
+of off-diagonal nonzeros in $A$. In the classical definition, the set
+of unknowns that unknown $i$ strongly depends upon is defined as
+\begin{equation}
+\label{2:eqn:S_i}
+S_i = \left\{j: i \ne j \textrm{ and }|a_{ij}| \ge \theta \max_{k \ne
+i}|a_{ik}|\right\},
+\end{equation}
+where $a_{ij}$ is the nonzero in row $i$, column $j$ of matrix $A$ and
+$0 < \theta \le 1$. The set of unknowns that $i$ strongly influences,
+denoted $S_i^T$, is defined as the set of unknowns that strongly
+depend on $i$: $S_i^T = \{j: i \in S_j\}$.
+
+This classical strength definition is formulated based on M-matrix
+properties and does not model the influence between unknowns in
+general cases. The formulation of new, effective, and inexpensive
+strength of connection measures is an active area of
+research~\cite{Brannick2005} and has the potential to extend the
+applicability of AMG solvers.
+
+Vertex $i$ is eligible to be added to the independent set $D$ if the
+following condition is satisfied:
+\begin{equation}
+  \label{eq:D-selection}
+  w_i > w_j,\, \forall j \in S_i \cup S_i^T.
+\end{equation}
 

### 2007-08-27 bsis_paper/paper.tex
* More in Section 2. * Some effort in Section 3.
git-svn-id: file:///home/rblake/svn/pamg/trunk@20 c7f3fe13-8c30-0410-9127-9a9c42fe009c

@@
 \label{sec:indep-set-based}
+Many independent set-based coarsening algorithms form $\Omega_{k+1}$
+using the following heuristics.
+\begin{description}
+\item[H1:] For each unknown $j$ that strongly influences $F$-point
+  $i$, $j$ is either a $C$-point or strongly depends on a $C$-point
+  $k$ that also strongly influences $i$.
+\item[H2:] The set of $C$-points needs to form a maximal independent
+  set in the reduced graph of $S$ such that no $C$-point strongly
+  depends on another $C$-point.
+\end{description}
+H1 and H2 cannot, in general, be satisfied simultaneously, so H1 is
+enforced by independent set-based methods, while H2 serves as a
+guideline whose purpose is to encourage selection of small
+$|\Omega_{k+1}|$.
+
+Some independent set-based coarsening algorithms use a variant of H1,
+called \hp, which is a relaxed version of H1.
+\begin{description}
+\item[\hp] Each $F$-point must be strongly influenced by at least one
+  $C$-point.
+\end{description}
+
 Independent set-based coarse grid selection algorithms implement three
@@
 
-Vertex $i$ is eligible to be added to the independent set $D$ if the
-following condition is satisfied:
+In classical coarsening, the independent set $D$ consists of a single
+vertex $i$ in each iteration, where $w_i \ge w_j,\, \forall j \in S_i
+\cup S_i^T$. Forming independent sets with only a single vertex is
+inefficient in parallel, so parallel independent set methods use
+different initialization techniques leading to the ability to form
+larger independent sets. In these algorithms vertex $i$ is eligible to
+be added to the independent set $D$ if the following condition is
+satisfied:
 \begin{equation}
@@
 \end{equation}
+The research presented here focuses on the independent set criterion
+in~(\ref{eq:D-selection}).
 
-Many independent set-based coarsening algorithms form $\Omega_{k+1}$
-using the following heuristics.
-\begin{description}
-\item[H1:] For each unknown $j$ that strongly influences $F$-point
-  $i$, $j$ is either a $C$-point or strongly depends on a $C$-point
-  $k$ that also strongly influences $i$.
-\item[H2:] The set of $C$-points needs to form a maximal independent
-  set in the reduced graph of $S$ such that no $C$-point strongly
-  depends on another $C$-point.
-\end{description}
-H1 and H2 cannot, in general, be satisfied simultaneously, so H1 is
-enforced by independent set-based methods, while H2 serves as a
-guideline whose purpose is to encourage selection of small
-$|\Omega_{k+1}|$.
+The weight update routines used in independent set-based coarsening
+algorithms are varied but primarily take one of two forms: routines
+that decrease the values of weights and routines that increase the
+value of weights. In all cases, a relatively large weight confers
+large value for having the corresponding vertex become a $C$-point.
 
-Some independent set-based coarsening algorithms use a variant of H1,
-called \hp, which is a relaxed version of H1.
-\begin{description}
-\item[\hp] Each $F$-point must be strongly influenced by at least one
-  $C$-point.
-\end{description}
+Many independent set-based coarsening algorithms have been
+developed. Ruge-St\"uben coarsening, also called classical coarsening,
+is introduced in~\cite{Ruge1987}. The first parallel independent
+set-based coarsening algorithms appeared approximately ten years ago
+and include RS3~\cite{Henson2002}, CLJP~\cite{Cleary1998}, BC-RS and
+Falgout~\cite{Henson2002}, PMIS and HMIS~\cite{Sterck2006},
+CLJP-c~\cite{alber-cljpc}, PMIS-c1 and PMIS-c2~\cite{alber-PCGS}, and
+PMIS Greedy~\cite{Butler2006}, to name a few.
 
@@
 considered thus far select independent sets using
-(\ref{5:eqn:D-conditions}), meaning $i$ is eligible to be in $D$ if
-its weight is larger than the weights of vertices in its neighborhood
-$\mathcal{N}_i$. Recall the neighborhood of $i$
+(\ref{eq:D-selection}), meaning vertex $i$ is eligible to be in $D$ if
+its weight is larger than the weights of vertices in its neighborhood,
+which is defined as follows.
+\begin{defn}
+\label{def:neighborhood}
+The \emph{neighborhood} of vertex $i$, denoted $\mathcal{N}_i$, is the
+union of the sets containing the strong influences of $i$ and the
+strong dependencies of $i$: $\mathcal{N}_i = S_i \cup S_i^T$.
+\end{defn}
+Recall the neighborhood of $i$
 (Definition~\ref{3:def:neighborhood}) is the set of vertices strongly

### 2007-09-18 bsis_paper/paper.tex
Edits and some new writing.
git-svn-id: file:///home/rblake/svn/pamg/trunk@24 c7f3fe13-8c30-0410-9127-9a9c42fe009c

@@
 finer level. Smoothed
-aggregation~\cite{Vanek1996,Vanek2001,brezina2005} forms a
-coarse-level degree of freedom by aggregating several fine-level
-degrees of freedom. In this paper, we study coarsening for classical
-AMG.
+aggregation~\cite{Vanek1996,Vanek2001,brezina2005}, on the other hand,
+forms a coarse-level degree of freedom by aggregating several
+fine-level degrees of freedom. In this paper, we study coarsening for
+classical AMG.
 
@@
 freedom $\Omega_{k+1} \subset \Omega_k$. Several approaches have been
-developed for selecting the coarse grid.
-
-Independent set-based algorithms are most numerous and all trace their
-roots to the classical AMG coarsening algorithm, RS
+developed for selecting the coarse grid, with independent set-based
+approaches the most numerous. Independent set-based algorithms trace
+their roots to the classical AMG coarsening algorithm, RS
 coarsening~\cite{Ruge1987}. These methods are our focus and are
@@
 Other methods take substantially different approaches to
-coarsening. The coarsening technique in~\cite{Griebel2006} develops a
-parallel coarsening algorithm that utilizes RS. On each processor, RS
-selects several different coarse grids, and then one coarse grid is
-selected per processor in a way to minimize special treatment on
-processor boundaries. In~\cite{MacLachlan2006}, a greedy approach is
-used to produce a good splitting of the unknowns into fine-grid and
+coarsening. The technique in~\cite{Griebel2006} develops a parallel
+coarsening algorithm that utilizes RS. On each processor, RS selects
+several different coarse grids, and then one coarse grid is selected
+per processor in a way to minimize special treatment on processor
+boundaries. In~\cite{MacLachlan2006}, a greedy approach is used to
+produce a good splitting of the unknowns into fine-grid and
 coarse-grid sets. Subdomain blocking techniques~\cite{Krechel2001}
@@
    $D \leftarrow\,$%\textsc{Select-Indep-Set}$(w,\, \Omega)$%
-   $(F,\, C,\, \Omega,\, w) \leftarrow\,$%\textsc{Update-Weights}$(F,\,
-                                C,\, \Omega,\, w)$%
+   $(F,\, C,\, \Omega,\, w) \leftarrow\,$%\textsc{Update-Weights}$(F,\,C,\, \Omega,\, w)$%
 end
@@
 becoming a $C$-point. The largest weights are conferred to vertices
-best suited to become $C$-points. The quality of potential $C$-points
-is quantified by \emph{strength of connection} measures. The classical
-strength of connection measure is based entirely on the relative sizes
-of off-diagonal nonzeros in $A$. In the classical definition, the set
-of unknowns that unknown $i$ strongly depends upon is defined as
+best suited to become $C$-points, where the quality of potential
+$C$-points is quantified by \emph{strength of connection}
+measures. The classical strength of connection measure is based
+entirely on the relative sizes of off-diagonal nonzeros in $A$. In the
+classical definition, the set of unknowns that unknown $i$ strongly
+depends upon is defined as
 \begin{equation}
@@
 inefficient in parallel, so parallel independent set methods use
-different initialization techniques leading to the ability to form
-larger independent sets. In these algorithms vertex $i$ is eligible to
-be added to the independent set $D$ if the following condition is
-satisfied:
+different initialization techniques designed to enable the selection
+of larger independent sets. In these algorithms, vertex $i$ is
+eligible to be added to the independent set $D$ if the following
+condition is satisfied:
 \begin{equation}
@@
 \end{equation}
-The research presented here focuses on the independent set criterion
-in~(\ref{eq:D-selection}).
+The research presented here focuses primarily on the independent set
+criterion in~(\ref{eq:D-selection}).
 
@@
 that decrease the values of weights and routines that increase the
-value of weights. In all cases, a relatively large weight confers
-large value for having the corresponding vertex become a $C$-point.
+value of weights. In all cases, a vertex with a large weight has high
+fitness for becoming a $C$-point.
 
@@
 \end{defn}
-Recall the neighborhood of $i$
-(Definition~\ref{3:def:neighborhood}) is the set of vertices strongly
-influenced by $i$ or strongly influencing $i$. Algorithms relying on
-different and larger neighborhoods, such as \emph{distance-$d$
-neighborhoods}, are conceivable.
+Algorithms that rely on different and larger neighborhoods, such as
+\emph{distance-$d$ neighborhoods}, are conceivable.
 \begin{defn}
@@
 this discussion assumes the weight of vertex $i$ is only decremented
-following weight updates and is never modified due to the assignment
-of a vertex $j \notin \mathcal{N}_i^s$ to the $C$-point set.
+following weight updates and is unaffected by the assignment of a
+vertex $j \notin \mathcal{N}_i^s$ to the $C$-point set.
 
@@
 Vertices in $D_0$ are destined to become $C$-points regardless of the
-algorithm used to build coarse grids. In the independent set-based
-algorithms discussed in Chapters~\ref{chap:CGS}
-and~\ref{chap:color-based}, all $D_0$ vertices become $C$-points in
+algorithm used to build coarse grids. In most independent set-based
+algorithms developed so far, all $D_0$ vertices become $C$-points in
 the first iteration. Any algorithm constructed, however, eventually

### 2007-09-19 bsis_paper/paper.tex
A few edits. Still some reorganization to do.
git-svn-id: file:///home/rblake/svn/pamg/trunk@26 c7f3fe13-8c30-0410-9127-9a9c42fe009c

@@
 \subsection{CLJP-c}
-Recall that CLJP-c colors the graph of $S$ before selecting
+CLJP-c~\cite{alber-cljpc} colors the graph of $S$ before selecting
 $C$-points, and the colors are used as one component of the vertex
@@
 \begin{cor}
-Any set of vertices in the graph of $S$ that share the same weight is
-an independent set.
+\label{cor:same-weight-indep-set}
+Any set of vertices in the graph of $S$ sharing the same weight are an
+independent set.
 \end{cor}
 \begin{cor}
-\label{5:cor:max-weight-indep-set}
+\label{cor:max-weight-indep-set}
 The set of vertices with the largest weight in the graph of $S$ form
@@
 \end{cor}
-The first corollary states that independent sets can be selected in
-the graph simply by selecting sets of vertices with the same
-weight. Corollary~\ref{5:cor:max-weight-indep-set} refines this
+Corollary~\ref{cor:same-weight-indep-set} states that independent sets
+can be selected in the graph simply by selecting sets of vertices with
+the same weight. Corollary~\ref{cor:max-weight-indep-set} refines this
 observation to a subset of vertices guaranteed to satisfy the
@@
 $C$-points and subsequently updating the weights of remaining vertices
-are developed. The new algorithm is labeled Bucket Sorted Independent
+are developed. The new algorithm is named Bucket Sorted Independent
 Sets (BSIS) to reflect the data structure used.

### 2007-09-20 bsis_paper/paper.tex
git-svn-id: file:///home/rblake/svn/pamg/trunk@27 c7f3fe13-8c30-0410-9127-9a9c42fe009c

@@
 \begin{abstract}
+Independent set-based coarse-grid selection algorithms are defined by
+policies regarding: weight initialization, independent set selection,
+and weight update. We develop theory demonstrating that all algorithms
+employing the same policies for each routine produce identical coarse
+grids, regardless of the implementation. This result on coarse grid
+invariance is applied to produce a new coarse-grid selection algorithm
+that is substantially more efficient than an existing algorithm that
+utilizes the same policies. Experimental results comparing two
+versions of the new algorithm and an older algorithm are presented,
+followed by a discussion on possible approaches to effectively
+parallelize the new algorithms.
 \end{abstract}
@@
 CLJP and its descendents select an independent set $D$ in each
-iteration. Stated originally in Section~\ref{3:sec:CLJP}, the
-condition for selecting $D$ is that all vertices $i \in D$ must
-satisfy
+iteration. Recall, the condition for selecting $D$ is that all
+vertices $i \in D$ must satisfy
 \begin{equation}
@@
 In each iteration, the non-empty bucket with largest weight forms $D$
-(see Corollary~\ref{5:cor:max-weight-indep-set}). These vertices are
+(see Corollary~\ref{cor:max-weight-indep-set}). These vertices are
 assigned to the $C$-point set and removed from the data

### 2007-09-23 bsis_paper/paper.tex
Major editing through the end of Section 3.
git-svn-id: file:///home/rblake/svn/pamg/trunk@28 c7f3fe13-8c30-0410-9127-9a9c42fe009c

--- 2007-09-19 bsis_paper/paper.tex
@@
 \begin{abstract}
+Independent set-based coarse-grid selection algorithms are defined by
+policies for how to do weight initialization, independent set
+selection, and weight update. In this paper, we develop theory
+demonstrating that all algorithms employing identical policies produce
+identical coarse grids, regardless of the implementation. This result
+on coarse grid invariance is applied to produce a new coarse-grid
+selection algorithm that is substantially more efficient than an
+existing algorithm using the same policies. Experimental results
+comparing two versions of the new algorithm and the older algorithm
+are presented, followed by a discussion on possible approaches to
+effectively parallelize the new methods.
 \end{abstract}
@@
 efficient numerical algorithm for finding the solution to numerous
-problems, including linear systems $Ax = b$ arising from the
-discretization of partial differential equations on structured and
+problems, including linear systems of the form $Ax = b$ arising from
+the discretization of partial differential equations on structured and
 unstructured meshes. In many cases, the computational complexity of
@@
 
-AMG algorithms are generally divided into two phases: the setup phase
-and the solve phase. Setup phase algorithms provide functionality for
-constructing elements needed by the solve phase, such as coarse grids,
-operators to transfer information between levels of the problem, and
-the linear systems to be solved on each level. The solve phase applies
-relaxation methods on the problems generated in the setup phase and
-uses the transfer operators to move vectors between levels.
+AMG algorithms execute in two phases: the setup phase and the solve
+phase. Setup phase algorithms provide functionality for constructing
+elements needed by the solve phase, such as coarse grids, operators to
+transfer information between levels of the problem (i.e., restriction
+and prolongation operators), and the linear systems to be solved on
+each level. The solve phase applies inexpensive iterative methods,
+called relaxation methods in the context of AMG, on the problems
+generated by the setup phase and uses the transfer operators to move
+vectors between levels.
 
 In this paper, we focus attention on coarse-grid selection algorithms
-in the AMG setup phase. One class of coarse-grid selection algorithms,
-independent set-based algorithms, is studied in detail. In
-Section~\ref{sec:cgs}, we introduce the coarse-grid selection problems
-and the goals of coarse-grid selection algorithms. The components of
-independent set-based coarsening algorithms are discussed.
+in the AMG setup phase. The independent set-based class of coarsening
+algorithms is studied in detail. In Section~\ref{sec:cgs}, we discuss
+current coarse-grid selection algorithms. The components of
+independent set-based coarsening algorithms are
+discussed. Section~\ref{sec:invariance} contains theory based on
+general independent set-based coarsening and the invariance of coarse
+grids selected when the same selection policies are
+used. Section~\ref{sec:bsis} introduces a new coarsening algorithm,
+Bucket-Sorted Independent Sets (BSIS), that, from the theory in
+Section~\ref{sec:invariance}, produces the same coarse grids as
+another coarsening algorithm in a much more efficient
+manner. Experimental results are reported to demonstrate the
+performance gains in BSIS discussion is included to propose a method
+for parallelization of BSIS. Conclusions are presented in
+Section~\ref{sec:conclusions}.
 
@@
 \label{sec:cgs}
-In AMG, coarse-grid selection is the process of selecting the degrees
-of freedom for the coarse level problem. In classical AMG, degrees of
-freedom on a coarse level are a subset of the degrees of freedom on a
-finer level. Smoothed
+Coarse-grid selection is the process of selecting the degrees of
+freedom for the coarse-level problem. Classical AMG uses nodal
+coarsening in which coarse-level degrees of freedom are a subset of
+the degrees of freedom on finer levels. Smoothed
 aggregation~\cite{Vanek1996,Vanek2001,brezina2005}, on the other hand,
 forms a coarse-level degree of freedom by aggregating several
-fine-level degrees of freedom. In this paper, we study coarsening for
-classical AMG.
+fine-level degrees of freedom. In this paper, we study nodal
+coarsening.
 
-Given a set of fine-level degrees of freedom, $\Omega_k$, coarse-grid
-selection for classical AMG selects a set of coarse-level degrees of
+Given a set of fine-level degrees of freedom, $\Omega_k$, nodal
+coarse-grid selection constructs a set of coarse-level degrees of
 freedom $\Omega_{k+1} \subset \Omega_k$. Several approaches have been
 developed for selecting the coarse grid, with independent set-based
-approaches the most numerous. Independent set-based algorithms trace
-their roots to the classical AMG coarsening algorithm, RS
-coarsening~\cite{Ruge1987}. These methods are our focus and are
-discussed in more detail in the next section.
+approaches being the most numerous. Independent set-based algorithms
+trace their roots to the classical AMG coarsening algorithm,
+Ruge-St\"uben (RS) coarsening~\cite{Ruge1987}, also called classical
+coarsening. These methods are our focus and are discussed in more
+detail following a brief overview of other techniques used for nodal
+coarsening.
 
@@
 guideline whose purpose is to encourage selection of small
-$|\Omega_{k+1}|$.
+coarse-grid point sets (i.e., small $|\Omega_{k+1}|$).
 
-Some independent set-based coarsening algorithms use a variant of H1,
-called \hp, which is a relaxed version of H1.
+Certain independent set-based coarsening algorithms use a variant of
+H1, called \hp, which is designed to lead to sparser coarse-level
+matrices.
 \begin{description}
@@
 
-Independent set-based coarse grid selection algorithms implement three
+Independent set-based coarse-grid selection algorithms implement three
 routines to select coarse grids: initialization, independent set
@@
    $D \leftarrow\,$%\textsc{Select-Indep-Set}$(w,\, \Omega)$%
-   $(F,\, C,\, \Omega,\, w) \leftarrow\,$%\textsc{Update-Weights}$(F,\,C,\, \Omega,\, w)$%
+   $C \leftarrow C \cup D$
+   $F \leftarrow F \setminus D$
+   $(\Omega,\, w) \leftarrow\,$%\textsc{Update-Weights}$(D,\, \Omega,\, w)$%
 end
@@
 becoming a $C$-point. The largest weights are conferred to vertices
-best suited to become $C$-points, where the quality of potential
-$C$-points is quantified by \emph{strength of connection}
-measures. The classical strength of connection measure is based
-entirely on the relative sizes of off-diagonal nonzeros in $A$. In the
-classical definition, the set of unknowns that unknown $i$ strongly
-depends upon is defined as
+best suited to be $C$-points, where the quality of potential
+$C$-points is quantified by \emph{strength of connection} measures. In
+the classical definition, the set of unknowns that unknown $i$
+strongly depends upon is based on the relative sizes of off-diagonal
+nonzeros in row $i$ of $A$ and is defined as
 \begin{equation}
@@
 denoted $S_i^T$, is defined as the set of unknowns that strongly
-depend on $i$: $S_i^T = \{j: i \in S_j\}$.
+depend on $i$: $S_i^T = \{j: i \in S_j\}$. The strong influence and
+strong dependence sets are used in the definition of the
+\emph{strength matrix} $S$, where the entry in row $i$, column $j$ of
+$S$ is defined as
+\begin{equation}
+S_{ij} = \left\{
+\begin{array}{ll}
+  1 & \textrm{if } j \in S_i, \\
+  0 & \textrm{otherwise}.
+\end{array}\right.
+\end{equation}
 
@@
 
-In classical coarsening, the independent set $D$ consists of a single
-vertex $i$ in each iteration, where $w_i \ge w_j,\, \forall j \in S_i
-\cup S_i^T$. Forming independent sets with only a single vertex is
-inefficient in parallel, so parallel independent set methods use
-different initialization techniques designed to enable the selection
-of larger independent sets. In these algorithms, vertex $i$ is
-eligible to be added to the independent set $D$ if the following
+The selection of the independent set $D$ works by comparing weights in
+each node's neighborhood, which is defined as follows.
+\begin{defn}
+\label{def:neighborhood}
+The \emph{neighborhood} of vertex $i$, denoted $\mathcal{N}_i$, is the
+union of the sets containing the strong influences of $i$ and the
+strong dependencies of $i$: $\mathcal{N}_i = S_i \cup S_i^T$.
+\end{defn}
+
+In classical coarsening, $D$ contains a single vertex $i$ in each
+iteration, where $w_i \ge w_j,\, \forall j \in
+\mathcal{N}_i$. Coarsening with independent sets containing a single
+vertex per iteration is not scalable, so parallel independent set
+methods use different initialization policies designed to enable the
+selection of larger independent sets. In these algorithms, vertex $i$
+is eligible to be added to the independent set $D$ if the following
 condition is satisfied:
@@
   \label{eq:D-selection}
-  w_i > w_j,\, \forall j \in S_i \cup S_i^T.
+  w_i > w_j\, \textrm{for all } j \in \mathcal{N}_i.
 \end{equation}
-The research presented here focuses primarily on the independent set
-criterion in~(\ref{eq:D-selection}).
 
-The weight update routines used in independent set-based coarsening
-algorithms are varied but primarily take one of two forms: routines
-that decrease the values of weights and routines that increase the
-value of weights. In all cases, a vertex with a large weight has high
-fitness for becoming a $C$-point.
+The weight update routines in independent set-based coarsening
+algorithms are varied but take one of two forms: routines that
+decrease the values of weights and routines that increase the value of
+weights. In all cases, a vertex with a large weight has high fitness
+for becoming a $C$-point.
 
 Many independent set-based coarsening algorithms have been
-developed. Ruge-St\"uben coarsening, also called classical coarsening,
-is introduced in~\cite{Ruge1987}. The first parallel independent
-set-based coarsening algorithms appeared approximately ten years ago
-and include RS3~\cite{Henson2002}, CLJP~\cite{Cleary1998}, BC-RS and
-Falgout~\cite{Henson2002}, PMIS and HMIS~\cite{Sterck2006},
+developed. RS coarsening is introduced in~\cite{Ruge1987} and is the
+inspiration for many algorithms that followed. The first parallel
+independent set-based coarsening algorithms appeared approximately ten
+years ago and include RS3~\cite{Henson2002}, CLJP~\cite{Cleary1998},
+BC-RS and Falgout~\cite{Henson2002}, PMIS and HMIS~\cite{Sterck2006},
 CLJP-c~\cite{alber-cljpc}, PMIS-c1 and PMIS-c2~\cite{alber-PCGS}, and
@@
 (\ref{eq:D-selection}), meaning vertex $i$ is eligible to be in $D$ if
-its weight is larger than the weights of vertices in its neighborhood,
-which is defined as follows.
-\begin{defn}
-\label{def:neighborhood}
-The \emph{neighborhood} of vertex $i$, denoted $\mathcal{N}_i$, is the
-union of the sets containing the strong influences of $i$ and the
-strong dependencies of $i$: $\mathcal{N}_i = S_i \cup S_i^T$.
-\end{defn}
-Algorithms that rely on different and larger neighborhoods, such as
+its weight is larger than the weights of all vertices in its
+neighborhood.
+
+Algorithms relying on different and larger neighborhoods, such as
 \emph{distance-$d$ neighborhoods}, are conceivable.
@@
 When a vertex $i$ satisfies the generalized selection
-condition~(\ref{5:eqn:general-selection}), no other $C$-point
-assignments affect $w_i$. This is formalized below.
+condition~(\ref{5:eqn:general-selection}), no other valid $C$-point
+assignments are able to affect $w_i$. This is formalized below.
 \begin{lem}
@@
 the first set of vertices satisfying the general selection conditions.
-\begin{defn}
+
 Let $D_0$ be the set of all vertices satisfying
 (\ref{5:eqn:general-selection}) in the first iteration of coarse-grid
-selection.
-\end{defn}
-
-Vertices in $D_0$ are destined to become $C$-points regardless of the
+selection. Vertices in $D_0$ become $C$-points regardless of the
 algorithm used to build coarse grids. In most independent set-based
@@
 \begin{proof}
-Let $c$ be the vertices in $\mathcal{N}_i^{2s}$ satisfying the
-conditions for $D$ in some arbitrary iteration. Suppose assigning
-vertices $c_1 \subset c$ to the $C$-point set leads to $w_i > w_j$ for
-all $j \in \mathcal{N}_i^s$. Also suppose assigning $c_2 \subset c$,
-$c_1 \ne c_2$, to the $C$-point set leads to the existence of some $j
-\in \mathcal{N}_i^s$ such that $w_j > w_k$ for all $k \in
-\mathcal{N}_j^s$.
+Let $c$ be the vertices in the extended selection neighborhood of $i$
+($\mathcal{N}_i^{2s}$) satisfying the conditions for $D$ in some
+arbitrary iteration. Suppose assigning vertices $c_1 \subset c$ to the
+$C$-point set leads to $w_i > w_j$ for all $j \in
+\mathcal{N}_i^s$. Also suppose assigning $c_2 \subset c$, $c_1 \ne
+c_2$, to the $C$-point set leads to the existence of some $j \in
+\mathcal{N}_i^s$ such that $w_j > w_k$ for all $k \in
+\mathcal{N}_j^s$. In the first case, $i$ becomes a $C$-point before
+$j$, and the opposite is true in the second case.
 
@@
 
-\begin{rem}
-CLJP and CLJP-c use the distance-one neighborhood as the selection
-neighborhood, (\ref{5:eqn:D-conditions}) as the selection criterion,
-and the weight update heuristics in
-Algorithm~\ref{3:alg:CLJP-update-heuristics}. Given the same initial
-weights, all algorithms based on the parameters utilized by CLJP and
-CLJP-c select identical coarse grids.
-\end{rem}
-
 Theorem~\ref{5:theorem:invariance} is an important result about the
@@
 \section{Bucket Sorted Independent Set Selection}
-\label{5:sec:contribution}
+\label{sec:bsis}
 
@@
 weight, and so on. This approach is taken by the algorithm developed
-in Section~\ref{5:sec:contribution}. It is proven in
+in Section~\ref{sec:bsissub}. It is proven in
 Section~\ref{5:sec:cg-invariance} that despite the difference in
@@
 \subsection{BSIS}
-\label{sec:bsis}
+\label{sec:bsissub}
 In this section, new techniques for searching the graph of $S$ for new

### 2007-09-24 bsis_paper/paper.tex
Overhauled Section 4, and wrote first content in the conclusions section (Section 5).
git-svn-id: file:///home/rblake/svn/pamg/trunk@29 c7f3fe13-8c30-0410-9127-9a9c42fe009c

@@
 \begin{equation}
-\label{5:eqn:general-selection}
+\label{eqn:general-selection}
 w_i > w_j\;\textrm{for all } j \in \mathcal{N}_i^s,
@@
 When a vertex $i$ satisfies the generalized selection
-condition~(\ref{5:eqn:general-selection}), no other valid $C$-point
+condition~(\ref{eqn:general-selection}), no other valid $C$-point
 assignments are able to affect $w_i$. This is formalized below.
@@
 \label{5:lemma:C-lock}
-If (\ref{5:eqn:general-selection}) is satisfied for vertex $i$, then
+If (\ref{eqn:general-selection}) is satisfied for vertex $i$, then
 $i$ must be the next vertex in $\{i\} \cup \mathcal{N}_i^s$ to become
@@
 \begin{proof}
-The satisfaction of (\ref{5:eqn:general-selection}) means $w_i > w_j$
+The satisfaction of (\ref{eqn:general-selection}) means $w_i > w_j$
 for all $j \in \mathcal{N}_i^s$. For $i$ to not become a $C$-point,
@@
 weight of $i$, however, is not decremented unless some $j \in
-\mathcal{N}_i^s$ satisfies~(\ref{5:eqn:general-selection}) and becomes
+\mathcal{N}_i^s$ satisfies~(\ref{eqn:general-selection}) and becomes
 a $C$-point, which is impossible until after $i$ is assigned to the
@@
 Let $D_0$ be the set of all vertices satisfying
-(\ref{5:eqn:general-selection}) in the first iteration of coarse-grid
+(\ref{eqn:general-selection}) in the first iteration of coarse-grid
 selection. Vertices in $D_0$ become $C$-points regardless of the
@@
 The proof follows from Lemma~\ref{5:lemma:C-lock}. All vertices in
-$D_0$ satisfy the selection condition~(\ref{5:eqn:general-selection}),
+$D_0$ satisfy the selection condition~(\ref{eqn:general-selection}),
 so the assignment of any other $C$-point has no effect on the weight
@@
 same initial weights and using the same selection neighborhood,
-selection criterion based on~(\ref{5:eqn:general-selection}), and
-weight update heuristics as described above select identical coarse
-grids.
+selection criterion based on~(\ref{eqn:general-selection}), and weight
+update heuristics as described above select identical coarse grids.
 \end{thm}
@@
 
-\section{Bucket Sorted Independent Set Selection}
+\section{Bucket-Sorted Independent Set Selection}
 \label{sec:bsis}
+In this section, we introduce Bucket-Sorted Independent Sets
+(BSIS). BSIS uses the same policies as CLJP-c, so from the theory in
+Section~\ref{sec:invariance}, BSIS and CLJP-c produce the same coarse
+grids. BSIS is designed to be more efficient than CLJP-c.
 
@@
 The set of vertices with the largest weight in the graph of $S$ form
-an independent set satisfying~\ref{5:eqn:D-conditions}. That is, each
-vertex in that independent set has a uniquely maximal weight in its
-neighborhood.
+an independent set satisfying~(\ref{eqn:general-selection}). That is,
+each vertex in that independent set has a uniquely maximal weight in
+its neighborhood.
 \end{cor}
@@
 selection criterion. In particular, it shows it is possible to build
-the coarse grid by selecting vertices with the maximum weight,
-updating weights, selecting the next set of vertices with maximum
-weight, and so on. This approach is taken by the algorithm developed
-in Section~\ref{sec:bsissub}. It is proven in
-Section~\ref{5:sec:cg-invariance} that despite the difference in
-approach, the resulting coarse grid is the same as that selected by
-CLJP-c.
+the coarse grid by selecting vertices with the maximum weight in the
+graph, updating weights, selecting the next set of vertices with
+maximum weight, and so on. This approach is taken by the BSIS
+algorithm.
 
-\subsection{BSIS}
-\label{sec:bsissub}
-In this section, new techniques for searching the graph of $S$ for new
-$C$-points and subsequently updating the weights of remaining vertices
-are developed. The new algorithm is named Bucket Sorted Independent
-Sets (BSIS) to reflect the data structure used.
-
-Like CLJP-c, BSIS depends on graph coloring, but utilizes modified
-routines for search and weight update. Furthermore, rather than
-applying the color information to augment vertex weights, BSIS uses
-the colors in a bucket data structure. Once initialized, this data
-structure selects independent sets, which satisfy the conditions in
-(\ref{5:eqn:D-conditions}), in constant time.
-
-\subsection{Motivation}
-\label{sec:motivation}
-CLJP and its descendents select an independent set $D$ in each
-iteration. Recall, the condition for selecting $D$ is that all
-vertices $i \in D$ must satisfy
-\begin{equation}
-\label{5:eqn:D-conditions}
-w_i > w_j\;\textrm{for all } j \in \mathcal{N}_i.
-\end{equation}
-CLJP and CLJP-c rely on a search routine to locate locally maximal
-weights and on a weight update routine to modify weights of vertices
-connected to new $C$-points.
+CLJP and CLJP-c rely on a search routine to locate all vertices
+satisfying~(\ref{eqn:general-selection}) (i.e., vertices with locally
+maximal weights) and on a weight update routine to modify weights of
+vertices in the neighborhoods of new $C$-points.
 
 The algorithms for searching and updating vertex weights in CLJP in
-detail are examined in this section. In particular, the impact of
-using a sparse matrix format on the coarsening procedure. The
-pseudo-code below assumes the software uses a compressed sparse row
-(CSR)~\cite{saadBook} matrix format or other similar format, which are
-common matrix formats in numerical software. CSR provides low memory
-costs for storing sparse matrices and provides efficient access to the
-nonzeros in a row. Accessing the nonzeros in a column is an expensive
-operation in this format and strongly influences the weight update
-routine in CLJP-style algorithms because $S$ is, in general, not
-symmetric.
+detail are examined in this section. The pseudo-code below assumes the
+software uses a compressed sparse row (CSR)~\cite{saadBook} matrix
+format or other similar format, which are common matrix formats in
+numerical software. CSR provides low memory costs for storing sparse
+matrices and provides efficient access to the nonzeros in a
+row. Accessing the nonzeros in a column is an expensive operation in
+this format and strongly influences the weight update routine in
+CLJP-style algorithms because $S$ is, in general, not symmetric.
 
@@
 $D \leftarrow \emptyset$
-for $i$ in $(C \cup F)$
-  if $w_i > w_j,\,\forall j \in \mathcal{N}_i$
+for $i \notin (C \cup F)$
+  if $w_i > w_j\ \textrm{for all } j \in \mathcal{N}_i$
     $D \leftarrow D \cup \{i\}$
@@
 
-Pseudo-code for the weight update in CLJP is shown in Algorithm~\ref{5:alg:setup}.
-\begin{lstlisting}[caption={CLJP Weight Update for CSR Matrix},label=5:alg:setup]
+Pseudo-code for the weight update in CLJP is shown in Algorithm~\ref{alg:setup}.
+\begin{lstlisting}[caption={CLJP Weight Update for CSR Matrix},label=alg:setup]
 %\textsc{Update-Weights}$(S,\, D,\, C,\, F,\, w)$%
@@
   for $i \in S_d$
-    $w_i \leftarrow w_i - 1$    //see %Figure~\ref{3:fig:CLJP-update-heuristics} left%
+    $w_i \leftarrow w_i - 1$
     $S_d = S_d \setminus \{i\}$ //removing edge from graph
@@
         if $k$ is marked NEW-C-POINT      // $i$ and $j$ mutually influenced by $k$
-          $w_j \leftarrow w_j - 1$        // see %Figure~\ref{3:fig:CLJP-update-heuristics} right%
+          $w_j \leftarrow w_j - 1$
           $S_i = S_i \setminus \{j\}$     // remove edge from $j$ to $i$
@@
 %\caption{CLJP Weight Update for CSR Matrix}
-%\label{5:alg:setup}
+%\label{alg:setup}
 %\textsc{Update-Weights}$(S,\, D,\, C,\, F,\, w)$ \{
@@
 $C$-points. When a new $C$-point $k$ is selected, the first type of
-weight update is trivial since determining the vertices in $S_k$ is
-inexpensive. The second type of update is more expensive since the
-vertices influenced by $k$ are difficult to determine in a CSR
-format. The update requires a search of many vertices $i$ and all of
-their strong influencing neighbors $j$. The routine then searches
-strongly influencing $j$ to determine if any $k \in D$ strongly
-influences both $i$ and $j$. The cost increases dramatically as the
-density of $S$ increases. Large operator complexities have a
-disproportionately large impact on coarse-grid selection run time. In
-the next section, a modified update routine to compliment the new
-search technique is introduced.
+weight update (Lines 2--7) is trivial since determining the vertices
+in $S_k$ is inexpensive. The second type of update (Lines 8--30) is
+more expensive since the vertices influenced by $k$ are difficult to
+determine in a CSR format. The update requires a search of many
+vertices $i$ and all of their strong influencing neighbors $j$. The
+routine then searches strongly influencing $j$ to determine if any $k
+\in D$ strongly influences both $i$ and $j$. The cost increases
+dramatically as the density of $S$ increases. Large operator
+complexities have a disproportionately large impact on coarse-grid
+selection run time. In the next section, a modified update routine to
+compliment the new search technique is introduced.
 
 \subsection{Bucket Sorted Independent Sets Algorithm}
-The BSIS algorithm creates a bucket data structure enabling a search
-for new $C$-points without individually scanning each vertex and
-associated edges. Figure~\ref{5:fig:bucket-algorithm} illustrates the
-bucket data structure. The number of buckets in the data structure is
-$\max_{i \in S}|S_i^T|$ times the number of colors in the graph. That
-is, each possible weight in $S$ has its own bucket. The vertices are
+\label{sec:bsis-alg}
+In this section, new techniques for searching the graph of $S$ for new
+$C$-points and subsequently updating the weights of remaining vertices
+are developed. The new algorithm is named Bucket Sorted Independent
+Sets (BSIS) to reflect the data structure used.
+
+Like CLJP-c, BSIS depends on graph coloring, but utilizes modified
+routines for search and weight update. Furthermore, rather than
+applying the color information to augment vertex weights, BSIS uses
+the colors in a bucket data structure. Once initialized, this data
+structure selects independent sets, which satisfy the conditions in
+(\ref{5:eqn:D-conditions}), in constant time.
+
+Figure~\ref{5:fig:bucket-algorithm} illustrates the bucket data
+structure. The number of buckets in the data structure is $\max_{i \in
+S}|S_i^T|$ times the number of colors in the graph. That is, each
+possible weight in $S$ has its own bucket. The vertices are
 distributed to the appropriate buckets during the setup of the
@@
 is black. Therefore, it is placed into the black bucket in the sixth
-group of buckets. More notably, the vertices in a bucket form an
-independent set (e.g., vertices 4, 16, and 21).
+group of buckets. Notably, the vertices in a bucket form an
+independent set (e.g., Vertices 14, 16, and 21).
 
@@
 \begin{lstlisting}[caption={BSIS Weight Update},label=5:alg:BSIS-weight-update]
-%\textsc{BSIS-Weight-Update}$(S)$%
+%\textsc{BSIS-Weight-Update}$(S,\, i)$%
 $bucketID \leftarrow (w_i-1)\cdot numColors + color_i$
@@
 structure following the first iteration of the algorithm. Vertex~10
-has become a $C$-point and its neighbors weights have been
+has become a $C$-point and its neighbors' weights have been
 updated. Vertices assigned to $F$ or $C$ are removed from the data
@@
     \includegraphics[width=0.98\textwidth]{images/BSIS/cp1}
-    \caption{The BSIS data structure after selecting the first
+    \caption{The BSIS data structure following selection of the first
     $C$-point (Vertex 10). The weights of neighbors of the new
@@
     other neighbors are moved to new locations in the data structure
-    and are highlighted (to help in reading the figure) in the data
+    and are highlighted (to aid in reading the figure) in the data
     structure with a red box.}
@@
 
-The weight update routine described in
-Section~\ref{5:sec:s-and-w-update} is very expensive in this context
-because some iterations of BSIS select few $C$-points. For a graph
-with a large number of colors, BSIS may execute dozens or hundreds of
-low-cost iterations to select a coarse grid. Recall the weight update
-routine described loops through all unassigned vertices each time it
-is called, so when it is run by BSIS, work is done on many unaffected
-vertices, which is computationally inefficient.
+The weight update routine described in Algorithm~\ref{alg:setup} is
+very expensive in this context because some iterations of BSIS select
+few $C$-points. For a graph with a large number of colors, BSIS may
+execute dozens or hundreds of low-cost iterations to select a coarse
+grid. Recall the weight update routine described loops through all
+unassigned vertices each time it is called, so when it is run by BSIS,
+work is done on many unaffected vertices.
 
@@
 searching for the second type of weight update in
-Algorithm~\ref{5:alg:setup}, which is done by looping through all
+Algorithm~\ref{alg:setup}, which is done by looping through all
 unassigned vertices since a new $C$-point cannot easily determine
@@
 remains approximately equivalent to the cost in
-CLJP-c. Chapter~\ref{chap:conclusions} describes plans for continued
-investigation into decreasing the cost of weight update in coarsening
-algorithms.
+CLJP-c.
 
@@
 
-Only one bucket in the data is touched during the $C$-point selection
-step: the largest weight non-empty bucket in the data structure. Other
-buckets are subsequently affected by the weight updates resulting from
-new $C$-point assignments. A different approach is possible, however,
+Only one bucket in the data structure is touched during the $C$-point
+selection step: the largest weight non-empty bucket. Other buckets are
+subsequently affected by the weight updates resulting from new
+$C$-point assignments. A different approach is possible, however,
 since the only bucket that must contain the correct vertices is the
@@
 \end{eqnarray}
-A 7-point Laplacian is selected since it is a common initial test
-problem and structured problems often lead to the largest operator
-complexities for coarsening algorithms satisfying heuristic H1. By
-creating larger operator complexities, the algorithms are forced to
-traverse more edges in the graph, leading to more work.
 
@@
 
-Experiment (\ref{5:eqn:Lap}) demonstrates the effectiveness and
-competitiveness of the bucket technique. Increased efficiency for the
-BSIS algorithm is anticipated through further research. Furthermore,
-the methods and concepts in this research are also applicable to other
-coarsening algorithms and possibly to other elements in the AMG setup
+The experiment demonstrates the effectiveness and competitiveness of
+the bucket technique. Increased efficiency for the BSIS algorithm is
+anticipated through further research. Furthermore, the methods and
+concepts in this research are also applicable to other coarsening
+algorithms, such as color-based methods designed to satisfy heuristic
+H1$^\prime$. The PMIS-c1 algorithm, for instance, is expected to
+benefit from BSIS and is expected to perform naturally in parallel
+since vertex weights are not updated in PMIS. The ideas in this
+research may also be applicable to other elements in the AMG setup
 phase.
-
-\subsection{BSIS Variants}
-The ideas developed in this chapter are applicable to other coarsening
-algorithms that utilize a graph coloring step, such as color-based
-methods designed to satisfy heuristic H1$^\prime$. The PMIS-c1
-algorithm is expected to benefit from BSIS and is expected to perform
-naturally in parallel since vertex weights are not updated in PMIS.
 
@@
 \label{sec:conclusions}
+In this paper, we have made two major contributions. First, new
+theoretical results were presented expanding knowledge on the nature
+of independent-set based coarse-grid selection. Algorithms using the
+same rules, with some restrictions, for weight initialization,
+independent set slection, and weight update produce the same coarse
+grids, regardless of the particular implementation of the
+rules. This result decouples the design of policies for selecting
+quality coarse grids from the design and efficiency of the algorithms
+used to select the coarse grids. The BSIS algorithm is designed as an
+example of this. BSIS selects the same coarse grids as CLJP-c, but
+uses a more efficient algorithm to select independent sets.
 

### 2007-09-24 bsis_paper/paper.tex
Typo fix.
git-svn-id: file:///home/rblake/svn/pamg/trunk@30 c7f3fe13-8c30-0410-9127-9a9c42fe009c

@@
 same rules, with some restrictions, for weight initialization,
-independent set slection, and weight update produce the same coarse
+independent set selection, and weight update produce the same coarse
 grids, regardless of the particular implementation of the

### 2007-10-01 bsis_paper/paper.tex
Small change to one proof.
git-svn-id: file:///home/rblake/svn/pamg/trunk@32 c7f3fe13-8c30-0410-9127-9a9c42fe009c
Additional changes to proof.
git-svn-id: file:///home/rblake/svn/pamg/trunk@33 c7f3fe13-8c30-0410-9127-9a9c42fe009c

@@
 \begin{proof}
-Let $c$ be the vertices in the extended selection neighborhood of $i$
-($\mathcal{N}_i^{2s}$) satisfying the conditions for $D$ in some
-arbitrary iteration. Suppose assigning vertices $c_1 \subset c$ to the
-$C$-point set leads to $w_i > w_j$ for all $j \in
-\mathcal{N}_i^s$. Also suppose assigning $c_2 \subset c$, $c_1 \ne
-c_2$, to the $C$-point set leads to the existence of some $j \in
-\mathcal{N}_i^s$ such that $w_j > w_k$ for all $k \in
-\mathcal{N}_j^s$. In the first case, $i$ becomes a $C$-point before
-$j$, and the opposite is true in the second case.
+  Let $c$ be the vertices in the extended selection neighborhood of
+  $i$ (i.e., $\mathcal{N}_i^{2s}$) satisfying the conditions for $D$
+  in some arbitrary iteration, where $i \notin c$. Suppose assigning
+  vertices $c_1 \subset c$ to the $C$-point set leads to $w_i > w_j$
+  for all $j \in \mathcal{N}_i^s$. Additionally, suppose assigning
+  $c_2 \subset c$, $c_1 \ne c_2$, to the $C$-point set leads to $w_j >
+  w_k$ for all $k \in \mathcal{N}_j^s$ for some $j \in
+  \mathcal{N}_i^s$, $j \notin c$. In the first case, $i$ becomes a
+  $C$-point before $j$, and the opposite is true in the second case.
 

### 2007-12-09 bsis_paper/paper.tex
Another editing pass. Changes in all sections. Updated bibliography file for articles published recently.
git-svn-id: file:///home/rblake/svn/pamg/trunk@40 c7f3fe13-8c30-0410-9127-9a9c42fe009c

@@
 Independent set-based coarse-grid selection algorithms are defined by
-policies for how to do weight initialization, independent set
-selection, and weight update. In this paper, we develop theory
-demonstrating that all algorithms employing identical policies produce
-identical coarse grids, regardless of the implementation. This result
-on coarse grid invariance is applied to produce a new coarse-grid
-selection algorithm that is substantially more efficient than an
-existing algorithm using the same policies. Experimental results
-comparing two versions of the new algorithm and the older algorithm
-are presented, followed by a discussion on possible approaches to
-effectively parallelize the new methods.
+policies for weight initialization, independent set selection, and
+weight update. In this paper, we develop theory demonstrating that
+algorithms employing the same policies produce identical coarse grids,
+regardless of the implementation. This result on coarse grid
+invariance is applied to produce a new coarse-grid selection algorithm
+that is substantially more efficient than an existing algorithm using
+the same policies. Experimental results comparing two versions of the
+new algorithm and the older algorithm are presented, followed by a
+discussion on possible approaches to effectively parallelize the new
+method.
 \end{abstract}
@@
 AMG algorithms execute in two phases: the setup phase and the solve
-phase. Setup phase algorithms provide functionality for constructing
+phase. Setup phase algorithms provide functions for constructing
 elements needed by the solve phase, such as coarse grids, operators to
@@
 in the AMG setup phase. The independent set-based class of coarsening
-algorithms is studied in detail. In Section~\ref{sec:cgs}, we discuss
-current coarse-grid selection algorithms. The components of
-independent set-based coarsening algorithms are
-discussed. Section~\ref{sec:invariance} contains theory based on
-general independent set-based coarsening and the invariance of coarse
-grids selected when the same selection policies are
-used. Section~\ref{sec:bsis} introduces a new coarsening algorithm,
-Bucket-Sorted Independent Sets (BSIS), that, from the theory in
-Section~\ref{sec:invariance}, produces the same coarse grids as
-another coarsening algorithm in a much more efficient
+algorithms is studied in detail. New theory is developed to
+demonstrate invariance in the result of independent set-based coarse
+grid selection, given restrictions on the selection
+policies. Algorithms that fall within the assumptions of the theory
+are guaranteed to produce the same coarse grids, regardless of how the
+policies are implemented. A new coarse-grid selection algorithm,
+designed to select coarse grids more efficiently than the CLJP-c
+algorithm, is presented to demonstrate the theoretical results. A
+particularly attractive result of the theory is that computational
+efficiency concerns become decoupled from the output of coarse-grid
+selection.
+
+The remainder of this paper is organized as follows. In
+Section~\ref{sec:cgs}, we discuss contemporary coarse-grid selection
+algorithms. The components of independent set-based coarsening
+algorithms are presented. Section~\ref{sec:invariance} develops theory
+based on general independent set-based coarsening and the invariance
+of coarse grids selected when the same selection policies are
+used. Section~\ref{sec:bsis} demonstrates the results from
+Section~\ref{sec:invariance} through the introduction of a new
+coarsening algorithm called Bucket-Sorted Independent Sets
+(BSIS). BSIS produces the same coarse grids as the CLJP-c coarsening
+algorithm, but is designed to do so in a much more efficient
 manner. Experimental results are reported to demonstrate the
-performance gains in BSIS discussion is included to propose a method
-for parallelization of BSIS. Conclusions are presented in
+performance gains in BSIS, and discussion is included to propose a
+method for the parallelization of BSIS. Conclusions are presented in
 Section~\ref{sec:conclusions}.
@@
 \label{sec:cgs}
-Coarse-grid selection is the process of selecting the degrees of
-freedom for the coarse-level problem. Classical AMG uses nodal
-coarsening in which coarse-level degrees of freedom are a subset of
-the degrees of freedom on finer levels. Smoothed
+Coarse-grid selection algorithms construct the degrees of freedom for
+the coarse-level problems. Classical AMG implements nodal coarsening
+in which coarse-level degrees of freedom form a subset of the degrees
+of freedom on finer levels. Smoothed
 aggregation~\cite{Vanek1996,Vanek2001,brezina2005}, on the other hand,
@@
 freedom $\Omega_{k+1} \subset \Omega_k$. Several approaches have been
-developed for selecting the coarse grid, with independent set-based
+developed to select the coarse grid, with independent set-based
 approaches being the most numerous. Independent set-based algorithms
-trace their roots to the classical AMG coarsening algorithm,
-Ruge-St\"uben (RS) coarsening~\cite{Ruge1987}, also called classical
-coarsening. These methods are our focus and are discussed in more
-detail following a brief overview of other techniques used for nodal
-coarsening.
+descend from the classical AMG coarsening algorithm, Ruge-St\"uben
+(RS) coarsening~\cite{Ruge1987}, also called classical
+coarsening. Independent set-based methods are our focus and are
+discussed in more detail in the next section.
 
@@
 approaches~\cite{Brandt2000,Livne2004,Brannick2005} form coarse grids
-incrementally and decide when the coarse grid is sufficient through a
-quality measure based on relaxation.
+incrementally and determine when the coarse grid is sufficient through
+a quality measure based on relaxation.
 
@@
 \label{sec:indep-set-based}
-Many independent set-based coarsening algorithms form $\Omega_{k+1}$
-using the following heuristics.
-\begin{description}
-\item[H1:] For each unknown $j$ that strongly influences $F$-point
-  $i$, $j$ is either a $C$-point or strongly depends on a $C$-point
-  $k$ that also strongly influences $i$.
-\item[H2:] The set of $C$-points needs to form a maximal independent
-  set in the reduced graph of $S$ such that no $C$-point strongly
-  depends on another $C$-point.
-\end{description}
-H1 and H2 cannot, in general, be satisfied simultaneously, so H1 is
-enforced by independent set-based methods, while H2 serves as a
-guideline whose purpose is to encourage selection of small
-coarse-grid point sets (i.e., small $|\Omega_{k+1}|$).
-
-Certain independent set-based coarsening algorithms use a variant of
-H1, called \hp, which is designed to lead to sparser coarse-level
-matrices.
-\begin{description}
-\item[\hp] Each $F$-point must be strongly influenced by at least one
-  $C$-point.
-\end{description}
-
-Independent set-based coarse-grid selection algorithms implement three
-routines to select coarse grids: initialization, independent set
-selection, and weight update. A general independent-set based
-coarsening algorithm is shown in
-Algorithm~\ref{alg:general-indep-set}.
-\begin{lstlisting}[caption={Independent Set-Based Coarse Grid
-Selection},label=alg:general-indep-set]
-%\textsc{Indep-Set-Coarse-Grid-Selection}$(\Omega)$:%
-$F \leftarrow \emptyset$
-$C \leftarrow \emptyset$
-$w \leftarrow\,$%\textsc{Initialize-Weights}$(\Omega)$%
-while $\Omega \ne \emptyset$
-   $D \leftarrow\,$%\textsc{Select-Indep-Set}$(w,\, \Omega)$%
-   $C \leftarrow C \cup D$
-   $F \leftarrow F \setminus D$
-   $(\Omega,\, w) \leftarrow\,$%\textsc{Update-Weights}$(D,\, \Omega,\, w)$%
-end
-\end{lstlisting}
-
-Each vertex in the graph is assigned a weight based on its merit to
-becoming a $C$-point. The largest weights are conferred to vertices
-best suited to be $C$-points, where the quality of potential
-$C$-points is quantified by \emph{strength of connection} measures. In
-the classical definition, the set of unknowns that unknown $i$
-strongly depends upon is based on the relative sizes of off-diagonal
-nonzeros in row $i$ of $A$ and is defined as
+Independent set-based coarsening algorithms rely on a \emph{strength
+  of connection} measure to estimate the level of correlation between
+  degrees of freedom. In the classical definition, the set of unknowns
+  that unknown $i$ strongly depends on is based on the relative sizes
+  of off-diagonal nonzeros in row $i$ of $A$ and is defined as
 \begin{equation}
@@
 
-The selection of the independent set $D$ works by comparing weights in
+Many independent set-based coarsening algorithms use strong influence
+and strong dependence to form the set of coarse-grid points
+$\Omega_{k+1}$ using the following heuristics.
+\begin{description}
+\item[H1:] For each unknown $j$ that strongly influences $F$-point
+  $i$, $j$ is either a $C$-point or strongly depends on a $C$-point
+  $k$ that also strongly influences $i$.
+\item[H2:] The set of $C$-points should form a maximal independent set
+  in the reduced graph of $S$ such that no $C$-point strongly depends
+  on another $C$-point.
+\end{description}
+H1 and H2 cannot, in general, be satisfied simultaneously, so H1 is
+enforced by independent set-based methods, while H2 serves as a
+guideline whose purpose is to encourage selection of small
+coarse-grid point sets (i.e., small $|\Omega_{k+1}|$).
+
+Certain independent set-based coarsening algorithms use a variant of
+H1, called \hp, which is designed to produce sparser coarse-level
+matrices.
+\begin{description}
+\item[\hp] Each $F$-point must be strongly influenced by at least one
+  $C$-point.
+\end{description}
+
+Independent set-based coarse-grid selection algorithms implement three
+routines to select coarse grids: initialization, independent set
+selection, and weight update. A general independent-set based
+coarsening algorithm is outlined in
+Algorithm~\ref{alg:general-indep-set}.
+\begin{lstlisting}[caption={Independent Set-Based Coarse Grid
+Selection},label=alg:general-indep-set]
+%\textsc{Indep-Set-Coarse-Grid-Selection}$(\Omega)$:%
+$F \leftarrow \emptyset$
+$C \leftarrow \emptyset$
+$w \leftarrow\,$%\textsc{Initialize-Weights}$(\Omega)$%
+while $\Omega \ne \emptyset$
+   $D \leftarrow\,$%\textsc{Select-Indep-Set}$(w,\, \Omega)$%
+   $C \leftarrow C \cup D$
+   $(F,\, w) \leftarrow\,$%\textsc{Update-Weights}$(D,\, F,\, \Omega,\, w)$%
+end
+\end{lstlisting}
+
+Each vertex in the graph is assigned a weight based on its merit for
+becoming a $C$-point. The largest weights are conferred to vertices
+best suited to be $C$-points, where the quality of potential
+$C$-points is quantified by strength of connection measures.
+
+The selection of the independent set $D$ in Line~6 of
+Algorithm~\ref{alg:general-indep-set} works by comparing weights in
 each node's neighborhood, which is defined as follows.
@@
 The weight update routines in independent set-based coarsening
-algorithms are varied but take one of two forms: routines that
-decrease the values of weights and routines that increase the value of
-weights. In all cases, a vertex with a large weight has high fitness
-for becoming a $C$-point.
+algorithms (encapsulated in Line~8 of
+Algorithm~\ref{alg:general-indep-set}) are varied but take one of two
+forms: routines that decrease the values of weights and routines that
+increase the value of weights. In all cases, a vertex with a large
+weight has high fitness and is more desirable as a $C$-point. As the
+pseudocode is formatted in Algorithm~\ref{alg:general-indep-set},
+assignment of vertices to the $F$-point set occurs during the weight
+update routine.
 
@@
 Algorithms relying on different and larger neighborhoods, such as
-\emph{distance-$d$ neighborhoods}, are conceivable.
+\emph{distance-$d$ neighborhoods}, are conceivable and are covered by
+the theory in this section.
 \begin{defn}
@@
 $\mathcal{N}_i^d$, is the set of vertices within $d$ hops of $i$ in
-the symmetrized strength matrix, excluding $i$. That is,
+the symmetrized strength graph, excluding $i$. That is,
 $\mathcal{N}_i^d = \left[\bigcup_{j \in
@@
 \label{5:lemma:C-lock}
-If (\ref{eqn:general-selection}) is satisfied for vertex $i$, then
-$i$ must be the next vertex in $\{i\} \cup \mathcal{N}_i^s$ to become
-a $C$-point, regardless of any other selections and updates made in
-the graph.
+If (\ref{eqn:general-selection}) is satisfied for vertex $i$, then $i$
+must be the next vertex in $\{i\} \cup \mathcal{N}_i^s$ to become a
+$C$-point, regardless of other $C$-point selections and corresponding
+weight updates made in the graph.
 \end{lem}
@@
 weight of $i$, however, is not decremented unless some $j \in
-\mathcal{N}_i^s$ satisfies~(\ref{eqn:general-selection}) and becomes
-a $C$-point, which is impossible until after $i$ is assigned to the
-$C$-point set.
+\mathcal{N}_i^s$ satisfies~(\ref{eqn:general-selection}) and becomes a
+$C$-point, which is impossible until $i$ is assigned to the $C$-point
+set.
 \end{proof}
@@
 Let $D_0$ be the set of all vertices satisfying
-(\ref{eqn:general-selection}) in the first iteration of coarse-grid
-selection. Vertices in $D_0$ become $C$-points regardless of the
+(\ref{eqn:general-selection}) following weight
+initialization. Vertices in $D_0$ become $C$-points regardless of the
 algorithm used to build coarse grids. In most independent set-based
@@
 method using the general selection condition invariably selects $D_0$
-vertices as $C$-points. This result is used in the proof of the next
-theorem.
+vertices as $C$-points. This result is used to prove, given the
+conditioned developed earlier in this section, the result of
+coarse-grid selection is insensitive to the implementation decisions
+made in specific algorithms.
 
@@
   Let $c$ be the vertices in the extended selection neighborhood of
-  $i$ (i.e., $\mathcal{N}_i^{2s}$) satisfying the conditions for $D$
-  in some arbitrary iteration, where $i \notin c$. Suppose assigning
-  vertices $c_1 \subset c$ to the $C$-point set leads to $w_i > w_j$
-  for all $j \in \mathcal{N}_i^s$. Additionally, suppose assigning
-  $c_2 \subset c$, $c_1 \ne c_2$, to the $C$-point set leads to $w_j >
-  w_k$ for all $k \in \mathcal{N}_j^s$ for some $j \in
+  $i$ (i.e., $ c \subset\mathcal{N}_i^{2s}$) satisfying the conditions
+  for $D$ in some arbitrary iteration. Suppose
+  assigning vertices $c_1 \subset c$ to the $C$-point set leads to
+  $w_i > w_j$ for all $j \in \mathcal{N}_i^s$. Additionally, suppose
+  assigning $c_2 \subset c$, $c_1 \ne c_2$, to the $C$-point set leads
+  to $w_j > w_k$ for all $k \in \mathcal{N}_j^s$ for some $j \in
   \mathcal{N}_i^s$, $j \notin c$. In the first case, $i$ becomes a
-  $C$-point before $j$, and the opposite is true in the second case.
+  $C$-point before $j$. The opposite is true in the second case.
 
@@
 the selection neighborhood of each vertex is invariant. Combined with
-Lemma~\ref{5:lemma:D0_invariant} as the base case, this proves by
-induction the invariance of coarse-grid selection for all algorithms
-using identical selection conditions.
+Lemma~\ref{5:lemma:D0_invariant} as the base case, induction proves
+the invariance of coarse-grid selection for all algorithms using
+identical selection and update conditions.
 \end{proof}
@@
 
-\section{Bucket-Sorted Independent Set Selection}
+\section{Bucket-Sorted Independent Sets Selection}
 \label{sec:bsis}
 In this section, we introduce Bucket-Sorted Independent Sets
-(BSIS). BSIS uses the same policies as CLJP-c, so from the theory in
+(BSIS). BSIS uses the same policies as the CLJP-c
+algorithm~\cite{alber-cljpc}, so from the theory in
 Section~\ref{sec:invariance}, BSIS and CLJP-c produce the same coarse
@@
 CLJP-c~\cite{alber-cljpc} colors the graph of $S$ before selecting
-$C$-points, and the colors are used as one component of the vertex
-weights. As a result, the structure of the coarse grids selected is
-improved. More formally, the use of graph coloring provides the
-following important result.
+$C$-points, where the colors are used as one component of the vertex
+weights and the number of strong influences provides the other source
+of contribution. As a result, the structure of the coarse grids
+selected is improved over the CLJP algorithm~\cite{Cleary1998}. More
+formally, the use of graph coloring provides the following important
+result.
 \begin{thm}
@@
 graph, updating weights, selecting the next set of vertices with
-maximum weight, and so on. This approach is taken by the BSIS
+maximum weight, and so on. This is the approach taken by the BSIS
 algorithm.
@@
 usually happens on the lower levels in the grid hierarchy
-(see~\cite{alber-PCGS} for examples). In the next section, a new
-technique is introduced for conducting the search in coarse-grid
-selection algorithms independent of the number of edges in the graph.
+(see~\cite{alber-PCGS} for examples). In Section~\ref{sec:bsis-alg},
+we introduce the technique used in BSIS, which conducts the search
+independent of the number of edges in the graph.
 
@@
 structure selects independent sets, which satisfy the conditions in
-(\ref{5:eqn:D-conditions}), in constant time.
+(\ref{eq:D-selection}), in constant time.
 
@@
 %\textsc{BSIS-Setup}$(S)$%
-for $i \in V$
+for $i \in S$
   $bucketID \leftarrow (w_i-1)\cdot numColors + color_i$
@@
 structure and other affected vertices are moved to new locations in
-the data structure. Such vertices are highlighted in red.
+the data structure. Such vertices are highlighted with a red ring.
 \begin{figure}
@@
 including the graph coloring phase. AMG solve phase information is not
-reported since the algorithms produce identical coarse grids and since
+reported since the algorithms produce identical coarse grids and
 information on solve phase performance for AMG with CLJP-c is
@@
 The smallest problem is a $30 \times 30 \times 30$ grid. Subsequent
-problems are grids of size $60^3$, $90^3$, up to $210^3$. The largest
-problem is 343 times larger than the smallest problem and contains
-more than nine million degrees of freedom.
+problems are grids of size $60^3$, $90^3$, and so on, up to
+$210^3$. The largest problem is 343 times larger than the smallest
+problem and contains more than nine million degrees of freedom.
 
@@
 replicate the operator matrix on processors, which leads to processors
-doing more of the same work, but by avoiding communication. The second
+doing some identical work, but by avoiding communication. The second
 and third ideas are similar in nature and both involve using dynamic
@@
 independent set selection, and weight update produce the same coarse
-grids, regardless of the particular implementation of the
-rules. This result decouples the design of policies for selecting
-quality coarse grids from the design and efficiency of the algorithms
-used to select the coarse grids. The BSIS algorithm is designed as an
-example of this. BSIS selects the same coarse grids as CLJP-c, but
-uses a more efficient algorithm to select independent sets.
+grids, regardless of the particular implementation of the rules. This
+result decouples the design of policies for selecting quality coarse
+grids from the design and efficiency of the algorithms used to select
+the coarse grids. Our second contribution demonstrates the power of
+the theoretical results with a new coarsening algorithm called
+BSIS. The BSIS algorithm is designed as an example of this. BSIS
+selects the same coarse grids as CLJP-c, but uses a more efficient
+algorithm to select independent sets.
 

### 2008-02-19 bsis_paper/paper.tex
changed some wordings and comment style
git-svn-id: file:///home/rblake/svn/pamg/trunk@42 c7f3fe13-8c30-0410-9127-9a9c42fe009c

@@
 
-\usepackage{amssymb,listings}
+\usepackage{amssymb,listings,color}
+\definecolor{mygray}{gray}{0.6}
 
@@
         language=Matlab,
+        morecomment=[n][\itshape\color{mygray}]{/*}{*/},
         belowcaptionskip=5pt}
@@
 
-\begin{abstract}
-Independent set-based coarse-grid selection algorithms are defined by
-policies for weight initialization, independent set selection, and
-weight update. In this paper, we develop theory demonstrating that
-algorithms employing the same policies produce identical coarse grids,
-regardless of the implementation. This result on coarse grid
-invariance is applied to produce a new coarse-grid selection algorithm
-that is substantially more efficient than an existing algorithm using
-the same policies. Experimental results comparing two versions of the
-new algorithm and the older algorithm are presented, followed by a
-discussion on possible approaches to effectively parallelize the new
-method.
+\begin{abstract} Independent set-based coarse-grid selection algorithms for
+algebraic multigrid are defined by their policies for weight initialization,
+independent set selection, and weight update. In this paper, we develop theory
+demonstrating that algorithms employing the same policies produce identical
+coarse grids, regardless of the implementation. The coarse grid invariance
+motivates a new coarse-grid selection algorithm that is more efficient than an
+existing algorithm using the same policies. Experimental results highlighting
+the efficiency of two versions of the new algorithm are presented, followed by a
+discussion on possible approaches to effectively parallelize the new method.
 \end{abstract}
@@
 The algebraic multigrid (AMG) method~\cite{Ruge1987,Brandt1982} is an
-efficient numerical algorithm for finding the solution to numerous
-problems, including linear systems of the form $Ax = b$ arising from
+efficient numerical algorithm for finding the solution to
+linear systems of the form $Ax = b$.  Often, these algebraic systems arise from
 the discretization of partial differential equations on structured and
@@
 AMG is $O(n)$, where $n$ is the number of unknowns in the linear
-system. The linear cost scaling property of AMG makes the method
-attractive, especially for large problems.
+system. The linear cost property of AMG makes the method particularly
+attractive for large scale problems.
 
@@
 algorithm, is presented to demonstrate the theoretical results. A
-particularly attractive result of the theory is that computational
-efficiency concerns become decoupled from the output of coarse-grid
-selection.
+particularly attractive result of the theory is that output from coarse-grid
+selection may be considered independent of the computational efficiency.
 
@@
 (BSIS). BSIS produces the same coarse grids as the CLJP-c coarsening
-algorithm, but is designed to do so in a much more efficient
+algorithm, but is designed to do so in a more efficient
 manner. Experimental results are reported to demonstrate the
@@
 (RS) coarsening~\cite{Ruge1987}, also called classical
-coarsening. Independent set-based methods are our focus and are
+coarsening. Independent set-based methods are the focus of this paper and are
 discussed in more detail in the next section.
@@
 
-Each vertex in the graph is assigned a weight based on its merit for
-becoming a $C$-point. The largest weights are conferred to vertices
+Each vertex in the graph is assigned a weight based on its merit in
+becoming a $C$-point. The largest weights are associated with vertices
 best suited to be $C$-points, where the quality of potential
@@
 The weight update routines in independent set-based coarsening
-algorithms (encapsulated in Line~8 of
-Algorithm~\ref{alg:general-indep-set}) are varied but take one of two
+algorithms (cf. Line~8 of
+Algorithm~\ref{alg:general-indep-set}) vary, but take one of two
 forms: routines that decrease the values of weights and routines that
 increase the value of weights. In all cases, a vertex with a large
-weight has high fitness and is more desirable as a $C$-point. As the
-pseudocode is formatted in Algorithm~\ref{alg:general-indep-set},
-assignment of vertices to the $F$-point set occurs during the weight
+weight has high fitness and is more desirable as a $C$-point. The
+pseudocode as written in Algorithm~\ref{alg:general-indep-set} performs
+assignment of vertices to the $F$-point set during the weight
 update routine.
@@
 Given a symmetric, but otherwise arbitrary, set of selection
-neighborhoods, the set of vertices potentially able to affect the
+neighborhoods, the set of vertices affecting the
 vertex weight of $i$ or $j \in \mathcal{N}_i^s$ is the \emph{extended
@@
 If (\ref{eqn:general-selection}) is satisfied for vertex $i$, then $i$
-must be the next vertex in $\{i\} \cup \mathcal{N}_i^s$ to become a
+is the next vertex in $\{i\} \cup \mathcal{N}_i^s$ to become a
 $C$-point, regardless of other $C$-point selections and corresponding
@@
 vertices as $C$-points. This result is used to prove, given the
-conditioned developed earlier in this section, the result of
+conditioned developed earlier in this section, that of
 coarse-grid selection is insensitive to the implementation decisions
@@
     $w_i \leftarrow w_i - 1$
-    $S_d = S_d \setminus \{i\}$ //removing edge from graph
+    $S_d = S_d \setminus \{i\}$ /*removing edge from graph*/
   end
@@
       for $k \in S_j$
-        if $k$ is marked NEW-C-POINT      // $i$ and $j$ mutually influenced by $k$
+        /* $i$ and $j$ mutually influenced by $k$ */
+        if $k$ is marked NEW-C-POINT      
           $w_j \leftarrow w_j - 1$
-          $S_i = S_i \setminus \{j\}$     // remove edge from $j$ to $i$
+          $S_i = S_i \setminus \{j\}$     /*remove edge from $j$ to $i$*/
         end
@@
       unmark $k$
-      $S_i = S_i \setminus \{k\}$         // remove edge from $k$ to $i$, if present
+      $S_i = S_i \setminus \{k\}$         /*remove edge from $k$ to $i$, if present*/
     end
